---
title: "Introduction to DATASCI 406"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
institute: "Computational Methods in Statistics and Data Science (DATASCI 406)"
date: ""
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: false
    includes:
      in_header: ../preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%",
                      fig.width = 6, fig.height = 2.4)
options(width = 60, digits = 4)
set.seed(49482)
library(tidyverse)
```

# Course Overview

## Course Goals

At the end of this course you will be able to
  
- Trade **computation time** for **analytical effort** 
- Connect **mathematical theory** to **computational applications** 
- Understand and explain **costs, benefits, and limits** of computational techniques 
- Implement key **statistical algorithms** in R 
- Create your own **statistical models**
- **Analyze data** using computational techniques
  
## What is Computational Statistics?

**Statistics**: the application of **probability theory** to real **data**. 
  
Goals:
  
- Specifying **models of data generation**
- Performing **inference** to connect real data to model
  - **Estimate** model parameters
  - **Test** hypotheses
  - **Predict** future observations
- Quantify **uncertainty** in the inference process.
- Understand **operating characteristics** of tools.
  
**Computational statistics** uses **substantial amounts of computation** to achieve these goals.

## Two fundamental concepts in computational statistics 

**Monte Carlo**: replace a quantity you cannot compute with an average over
random draws. If we can get a sample $X_1, \ldots, X_n$, then
 

$$\E{g(X)} = \int g(x) f(x) \, dx \;\approx\; \frac{1}{n} \sum_{i=1}^n g(X_i)$$

**Optimization**: search for the parameter value that makes something best,
when you cannot solve for it algebraically. 

E.g., for a complicated density $f$ with parameter $\theta$, what value of $\theta$ would be most consistent with my sample?

## Four Modules

| Module | Name | Topic |
|---|---|---|
| 1 | **Integrate** | Monte Carlo with the distributions you have |
| 2 | **Generate** | Monte Carlo when you need a new distribution |
| 3 | **Resample** | Monte Carlo with only your data |
| 4 | **Fit** | Optimization |

Let's see some examples!

# Module 1: Integrate

## A density

Consider a **random variable** $X$ with density $f$:

$$f(x) = \frac{1}{\sqrt{2\pi}}
  \exp\left\{ 1.5\, x - 0.25\, x^2 - 2.5966 \right\}$$

How would you find $\E{X} = \int_{-\infty}^{\infty} x f(x) \, dx$?

Nothing here looks like an integral you can do by hand.

## Draw, then average

```{r echo=FALSE}
sampleX <- function(n) rnorm(n, mean = 3, sd = sqrt(2))
```

Suppose I hand you `sampleX(n)`, which returns $n$ draws from $f$.

```{r}
x <- sampleX(100000)
mean(x)
```

## More complicated integrals

\[
\begin{aligned}
\E{|X|}              &= \int_{-\infty}^{\infty} |x|\, f(x)\, dx \\[0.3em]
\Pr(X > 5)           &= \int_{5}^{\infty} f(x)\, dx \\[0.3em]
\E{\log(1 + e^X)}    &= \int_{-\infty}^{\infty} \log(1 + e^x)\, f(x)\, dx \\[0.3em]
\E{(1 + X^2)^{-1}}   &= \int_{-\infty}^{\infty} \frac{f(x)}{1 + x^2}\, dx
\end{aligned}
\]

## Same code, different $g$

Each is $\E{g(X)} = \int g(x) f(x)\, dx$ for a different $g$.

```{r}
c(EabsX  = mean(abs(x)),
  P_gt_5 = mean(x > 5),
  Esoft  = mean(log1p(exp(x))),
  Einv   = mean(1 / (1 + x^2)))
```

Change $g$, keep everything else.

## Sampling in R

Every common distribution has an `r` function:

| | | | |
|---|---|---|---|
| `rnorm` | Normal | `rbinom` | Binomial |
| `runif` | Uniform | `rpois` | Poisson |
| `rexp` | Exponential | `rgamma` | Gamma |
| `rt` | Student's $t$ | `rbeta` | Beta |

All take the number of draws first: `rnorm(n, mean, sd)`.

```{r}
sampleX
```
## Functions of samples

A **statistic** is a function of a sample:

$$T = T(X_1, X_2, \ldots, X_n)$$

We use statistics to

- Estimate population parameters
- Test hypotheses about populations
- Perform prediction for new observations

A statistic is a random variable, so it has a distribution.

## Distributions of statistics 

$$T = \max_{i = 1,2,3} X_i$$

```{r}
sample_max <- function(n) max(sampleX(n))
empirical_distrib <- replicate(100000, sample_max(3))
```

##

```{r echo=FALSE}
ggplot(data.frame(x = empirical_distrib), aes(x = x)) +
  geom_histogram(bins = 60) +
  labs(x = "max of 3 draws", y = NULL)
```

## Some Properties

```{r }
mean(empirical_distrib); var(empirical_distrib)
```

```{r }
quantile(empirical_distrib, c(0.25, 0.5, 0.75))
```

These are **estimates**. How far off? That is module 1.

# Module 2: Generate

## A distribution R does not have

The **Pareto** distribution: incomes, city sizes, file sizes. A heavy right
tail, so extreme values are far more common than under a Normal.

For shape $\alpha > 0$ and $x \ge 1$:

$$\Pr(X > x) = x^{-\alpha}$$
Therefore it has a **cumulative distributiuon function** (CDF)

$$F(x) = 1 - x^{-\alpha}, \qquad x \ge 1$$

## The quantile function

Given a probability $u$, which $x$ has $F(x) = u$?

$$u = 1 - x^{-\alpha}
\quad\Longrightarrow\quad
x^{-\alpha} = 1 - u
\quad\Longrightarrow\quad
F^{-1}(u) = (1 - u)^{-1/\alpha}$$

$F^{-1}(0.5)$ is the median, $F^{-1}(0.9)$ the 90th percentile.

```{r}
qpareto <- function(u, alpha = 3) (1 - u)^(-1/alpha)
qpareto(c(0.5, 0.9, 0.99))
```

## How to sample? Inversion!

Later we will prove: If $U \sim \text{Unif}(0,1)$, then $F^{-1}(U)$ has CDF $F$.

Algorithm in R:

- Sample $U$ using `runif`
- Create $X$ using $F^{-1}(U)$

## A Pareto sampler in three lines

```{r}
rpareto <- function(n, alpha = 3) qpareto(runif(n), alpha)

x <- rpareto(100000)
c(mean = mean(x), exact = 3 / 2, min = min(x))
```
##

```{r echo=FALSE}
ggplot(data.frame(x = x[x < 5]), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60) +
  #stat_function(fun = function(z) alpha * z^(-alpha - 1), colour = "red") +
  labs(x = "x (truncated at 5 for display)", y = NULL)
```

# Module 3: Resample

## Unknown distributions

In the previous examples, we had **known distributions** (later we will see these in hypothesis tests and estimating parameters of assumed population distributions).

What about situations where we **don't want to assume a population distribution?**

Idea: Instead of writing down a CDF $F$, what if we estimated it? 

## What distribution is this?

Systolic and diastolic blood pressure, NHANES 2013--14.

```{r}
load("../../data/aspirin.rda")
dim(combined)
```

```{r echo=FALSE, fig.height = 2.1}
ggplot(combined, aes(x = sys_mean, y = dia_mean)) +
  geom_point(alpha = 0.08) +
  labs(x = "systolic", y = "diastolic")
```

## No named family fits

```{r echo=FALSE}
skew <- function(x) mean(((x - mean(x)) / sd(x))^3)
c(systolic = skew(combined$sys_mean), diastolic = skew(combined$dia_mean))
```

The two margins skew in **opposite directions**, so this is not bivariate
Normal. There is also a clump of implausibly low diastolic readings.

We could go hunting for a family that fits. Or we could stop assuming one.

## Estimating the CDF

The **empirical CDF** puts mass $1/n$ on each observed value:

$$\hat F(x) = \frac{1}{n} \sum_{i=1}^{n} \mathbf{1}\{X_i \le x\}$$

```{r echo=FALSE, fig.height = 2}
set.seed(406)
sub <- sample(combined$sys_mean, 40)
ggplot(data.frame(x = sub), aes(x = x)) +
  stat_ecdf(geom = "step") +
  stat_function(fun = pnorm, args = list(mean = mean(combined$sys_mean),
                                         sd = sd(combined$sys_mean)),
                colour = "red") +
  labs(x = "systolic BP", y = expression(hat(F)(x)))
```

## Inversion, one module later: The Bootstrap

Module 2: to sample from $F$, apply $F^{-1}$ to a uniform draw.

Now do exactly that to $\hat F$. Since $\hat F$ puts mass $1/n$ on each
observed point, $\hat F^{-1}(U)$ returns **one of your data points, chosen
uniformly at random.**

$$\text{sampling from } \hat F \;=\; \texttt{sample(x, replace = TRUE)}$$

We call this ``bootstrapping.'' It is inversion applied to the CDF you estimated.

## Correlation of systolic and diastolic BP

Blood pressure, NHANES 2013--14:

```{r}
theta <- cor(combined$sys_mean, combined$dia_mean)
c(n = nrow(combined), r = theta)
```

What is the **standard error of that correlation**? What would be a good **95% confidence interval?**

## Bootstrapping for SE, quantiles

Treat the sample as the population and draw from it.

```{r}
set.seed(406)
boot_r <- replicate(4000, {
    i <- sample(nrow(combined), replace = TRUE)   # rows, not values
    cor(combined$sys_mean[i], combined$dia_mean[i])
})
c(se = sd(boot_r), quantile(boot_r, c(0.025, 0.975)))
```

##

```{r echo=FALSE}
ggplot(data.frame(r = boot_r), aes(x = r)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = theta, colour = "red") +
  labs(x = "bootstrap correlation", y = NULL)
```

# Module 4: Fit

## Only the maximum survives

Same $X \sim f$ as before, $\sigma^2 = 2$ known. We draw $m = 200$ groups of
three, but the record keeps only the largest value in each group:

$$M_j = \max\left(X_{j1},\, X_{j2},\, X_{j3}\right), \qquad j = 1, \ldots, 200$$

```{r}
set.seed(406)
m <- 200
groups <- matrix(sampleX(m * 3), nrow = m)
M <- apply(groups, 1, max)   # all we get to see
```

Estimate $\mu$.

## How to estimate?

Note that $\mu$ is not the same as $\E{\max X}$:

```{r}
mean(M)
```

1. Maximum likelihood estimation
2. Algorithm that converges to the MLE

## Maximum likelihood, briefly

The **likelihood** is the probability of the data we saw, read as a function of
the unknown parameter:

$$L(\mu) = \prod_{j=1}^{m} f_M(M_j; \mu)
\qquad\qquad
\hat\mu = \arg\max_\mu \; L(\mu)$$

Pick the $\mu$ that makes what we observed as likely as possible. More on this
in the statistical review.

## Why not just maximize the likelihood?

The max of three draws has density $3 f(x) F(x)^2$, so

$$\ell(\mu) = \sum_{j=1}^{m}
  \left[ \log\phi\!\left(\tfrac{M_j - \mu}{\sigma}\right)
       + 2\log\Phi\!\left(\tfrac{M_j - \mu}{\sigma}\right) \right]$$

Setting $\ell'(\mu) = 0$ gives

$$\sum_{j=1}^{m} \frac{M_j - \mu}{\sigma^2}
  \;=\; \frac{2}{\sigma} \sum_{j=1}^{m}
  \frac{\phi\!\left(\tfrac{M_j-\mu}{\sigma}\right)}
       {\Phi\!\left(\tfrac{M_j-\mu}{\sigma}\right)}$$

$\mu$ is stuck inside $\Phi$, which is itself an integral with no closed form.
**You cannot solve this for $\mu$.**

## What we would do with the full data

With all $3m$ values in hand, estimating $\mu$ is a sample mean:

$$\hat\mu = \frac{1}{3m} \sum_{j=1}^{m} \underbrace{\sum_{i=1}^{3} X_{ji}}_{\textstyle S_j}$$

We only need the **group totals** $S_j$, not the individual values.

We cannot compute $S_j$. But we know one of its three terms and have a bound on the other two.

## Making a reasonable guess

Replace the total we cannot compute with **its expectation given what we saw.**

Given $M_j$, one value equals $M_j$ and the other two are draws from $f$
conditioned to lie below $M_j$:

$$\E{S_j \mid M_j} \;=\; M_j \;+\; 2\,\E{X \mid X \le M_j}$$

For a Normal the conditional mean is closed form:

$$\E{X \mid X \le a} \;=\; \mu - \sigma\,
  \frac{\phi\!\left(\frac{a-\mu}{\sigma}\right)}
       {\Phi\!\left(\frac{a-\mu}{\sigma}\right)}
  \;<\; \mu$$

It depends on $\mu$, which is what we are trying to find. So guess, then iterate.

## Example

Start from the observed average maximum, $\mu^{(0)} = 4.30$, and take the first
group, which came back with $M_j = 5.77$:

$$\E{X \mid X \le 5.77} = 3.91$$

$$\E{S_j \mid M_j} = 5.77 + 2(3.91) = 13.60$$

The true total was $1.16 + 2.57 + 5.77 = 9.50$.

One group is badly wrong. The M step averages 200 of them.

## EM: Expectation-Maximization algorithm

Start with a guess $\mu^{(0)}$. For $t = 1, 2, \ldots$:

1. **E**: compute $\E{S_j \mid M_j}$ for every group, using $\mu^{(t-1)}$.
2. **M**: set $\mu^{(t)} = \frac{1}{3m}\sum_j \E{S_j \mid M_j}$.
3. Repeat until it **convergence** (estimate stops moving).

Starting from $\mu^{(0)} = 4.30$, one step gives $3.49$.

## EM in R

```{r}
sigma <- sqrt(2)
Etrunc <- function(a, mu, s) mu - s * dnorm((a-mu)/s) / pnorm((a-mu)/s)

guess <- mean(M)
for (i in 1:100) {
    e_step <- (M + 2 * Etrunc(M, guess, sigma)) / 3
    m_step <- mean(e_step)
    guess <- m_step
}
c(EM = m_step, naive = mean(M), mu = 3)
```

## Convergence

```{r echo=FALSE}
g <- mean(M); tr <- g
for (i in 1:25) { g <- mean((M + 2 * Etrunc(M, g, sigma)) / 3); tr <- c(tr, g) }
mle <- optimize(function(mu) sum(dnorm(M, mu, sigma, log = TRUE) +
                                 2 * pnorm(M, mu, sigma, log.p = TRUE)),
                c(0, 8), maximum = TRUE)$maximum

ggplot(data.frame(step = seq_along(tr) - 1, mu = tr), aes(x = step, y = mu)) +
  geom_hline(yintercept = mle, colour = "red") +
  geom_hline(yintercept = 3, colour = "grey50", linetype = "dashed") +
  geom_line() + geom_point(size = 1.3) +
  labs(x = "iteration", y = expression(mu^(t)))
```

Red is the **MLE from the data we actually have**; dashed grey is $\mu = 3$.

## What is EM converging to?

Not $\mu$, and not the average of the values we lost. EM maximizes the
likelihood of the data **we observed**:

| | |
|---|---:|
| population $\mu$ | 3.000 |
| mean of the 600 hidden values | 3.063 |
| **MLE from the 200 maxima** | **`r round(mle, 3)`** |
| EM after 25 steps | `r round(tr[length(tr)], 3)` |

EM lands on the MLE, exactly. The gap from 3 is **sampling error in 200
maxima**, not a flaw in the algorithm. With more groups the MLE moves toward 3.

# Course Logistics

## People

\begin{description}[Student Hours]
\item[Professor] Mark Fredrickson (<mfredric@umich.edu>)
\item[GSIs] Ishan Paul (<ishanpl@umich.edu>), Qingyuan Yu (<qyyu@umich.edu>)
\item[Student Hours] TBA, see Canvas
\end{description}

## Prerequisites

   
- Some programming experience (R, Python, C/C++, Java)
- An understanding of the following terms: Distributions and random variables, random sampling and sampling distributions, population parameters and estimation, hypothesis tests, basic calculus (integrals, derivatives).
   
## Course Materials

No required text, but several **recommended** (particularly first 2). Links to electronic copies available through the library on Canvas.
  
- Rizzo, Maria L. *Statistical Computing with R, 2ed*
- Wickham, H. \& Grolemund, G. *R for Data Science*
- Irizarry, R. A. *Introduction to Data Science*
- Agresti, A. *Foundations of Linear and Generalized Linear Models*
- Gentle, J. E. *Computational Statistics*
- *Handbook of Computational Statistics*. H\"{a}rdle, Gentle, and Mori, eds.
  
All slides and videos will be posted to **Canvas** and **OneNote**.

## Approach

The class is composed of **four modules** and uses a **flipped classroom** approach.

Each module consists of short **check-in quizzes**, a **project**, and an **exam**. There is no separate final exam.

Materials are presented in **video lectures**. Class time is for short **check-in quizzes**, working on examples and the project, and exams.

## Quizzes

- 4--6 multiple choice plus one short answer, every session
- Covers the material you watched **before** class
- **Lowest per module dropped. No makeups.**
- I will reach out to students with testing accommodations.

## Projects

Each module includes a project. The project gives you an opportunity to:

- Implement key statistical algorithms
- Investigate the properties of those algorithms
- Apply these techniques to real-world data

These projects become your notes for your exams. **You will be asked questions on the exams about your project.**

## Exams

Each module concludes with an exam. Approximate coverage:

- 50\%: General conceptual questions, similar to the quizzes.
- 50\%: Questions about the results of your own project.

You may bring any **paper** notes you wish. No electronic devices. Held during class periods, with the exception of the Module 4 exam, which uses the scheduled final exam time.

Students with accommodations will be contacted with alternate exam location information.

## Grading 
 
Each module counts for 25% of your grade. Within each module:

- Exam: 20% of overall grade
- Quizzes: 5% of overrall grade, one drop per module

## AI policy

> **Permitted and expected** on projects and all out-of-class work.
>
> **Not permitted** on quizzes or exams.

Remember: you will be asked about project details. Be prepared to explain coding details, even if you used AI to assist.

## Other Policies

- Course Recordings
- Disability Statement
- Mental Health and Well-Being
- Sexual Misconduct Policy

## Tour of Canvas

## Next Session

- No quiz
- Video: statistical review
- Class period: Introduction to R and RMarkdown
- All students: install R and RStudio on laptops, directions on Canvas

