---
pagetitle: "Quiz 2.1: The Inverse Transform"
output:
  pdf_document: default
geometry: "margin=0.75in, top=0.6in, bottom=0.6in"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
```

<!-- Side-by-side columns. Macros rather than a bare minipage, so pandoc
     still reads the markdown between them. -->
\newcommand{\colstart}[1]{\noindent\begin{minipage}[c]{#1\textwidth}}
\newcommand{\colnext}[1]{\end{minipage}\hfill\begin{minipage}[c]{#1\textwidth}}
\newcommand{\colend}{\end{minipage}}

**DATASCI 406 FA 26, Quiz 2.1**

\vspace{0.3em}
\noindent
\begin{tabular*}{\textwidth}{|@{\extracolsep{\fill}}p{0.40\textwidth}|p{0.22\textwidth}|p{0.22\textwidth}|}
\hline
\small Name & \small Uniqname & \small UMID \\[3.6em]
\hline
\end{tabular*}
\vspace{0.3em}

Each question is worth 1 point. Circle one answer for each.

## Question 1 (1 point)

For a random variable $X$ and a probability $p$, what does the quantile
function $Q_X(p)$ return?

1. The probability that $X$ is less than or equal to $p$.
1. The smallest value $x$ such that the CDF evaluated at $x$ is at least $p$.
1. The value $x$ at which the density of $X$ equals $p$.
1. The value of the CDF evaluated at $X$.

## Question 2 (1 point)

$X$ has CDF $F(x) = x^3$ for $0 \le x \le 1$. What is its quantile function
$Q(u)$ for $0 \le u \le 1$?

\noindent\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}llll@{}}
1. $u^3$ & 2. $3u^2$ & 3. $1 - u^{1/3}$ & 4. $u^{1/3}$
\end{tabular*}

## Question 3 (1 point)

Which line of R produces 1000 draws from an exponential distribution with
rate 2, using `qexp()` and not `rexp()`?

1. `runif(1000, qexp(0, rate = 2), qexp(1, rate = 2))`
1. `qexp(1000, rate = 2) * runif(1000)`
1. `qexp(runif(1000), rate = 2)`
1. `runif(qexp(1000, rate = 2))`

## Question 4 (1 point)

\colstart{0.56}

A sampler is meant to draw from $\text{Exp}(1)$. The QQ plot at right
compares 500 of its draws with the $\text{Exp}(1)$ quantiles, and the line is
the 45 degree line. Is the sampler correct?

1. Correct: the points form a straight line.
1. Wrong: the points do not follow the 45 degree line.
1. Can't tell: a plot is not a hypothesis test.

\colnext{0.40}

```{r fig.width=3.2, fig.height=2.4, out.width="100%"}
set.seed(2261)
x <- -2 * log(runif(500))
ggplot(data.frame(sample = x), aes(sample = sample)) +
  geom_qq(distribution = qexp, size = 0.6) +
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "Exp(1) quantiles", y = "Sample quantiles") +
  theme_bw(base_size = 9)
```

\colend

## Question 5 (1 point)

$X$ takes the values below. To draw $X$ you generate $u = 0.53$ and apply
the inverse transform method for discrete variables. What value $x$ do you generate?

| $x$ | 10 | 20 | 30 | 40 |
|---|:-:|:-:|:-:|:-:|
| $\Pr(X = x)$ | 0.2 | 0.3 | 0.1 | 0.4 |

\noindent\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}llll@{}}
1. 10 & 2. 20 & 3. 30 & 4. 40
\end{tabular*}

