---
title: "The Inverse Transform: Recap"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
institute: "Computational Methods in Statistics and Data Science (DATASCI 406)"
date: "Module 2: Generate, Session 1"
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: false
    includes:
      in_header: ../preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = "100%")
library(ggplot2)
```

## Pseudorandom numbers

A **PRNG** maps a seed to a number and a new seed:
$$\text{seed}_k \;\longmapsto\; (U_k, \text{seed}_{k+1})$$

- **Deterministic** (not inherently true randomness)
- The **seed** makes a simulation **reproducible**: `set.seed()`.
- It **repeats** eventually: the LCG after 8, Mersenne-Twister after $2^{19937} - 1$.
- We cannot **prove** a generator is good, only **reject** bad ones.

We will use `runif()` as our PRNG.

## The quantile function

For any random variable, with CDF $F_X(t) = \Pr(X \le t)$:
$$Q_X(p) = \inf\left\{x : F_X(x) \ge p\right\}, \qquad p \in [0, 1]$$

- "The **smallest** $x$ whose CDF reaches $p$."
- Defined for **every** distribution: continuous, discrete, or mixed.
- When $F_X$ is continuous and strictly increasing, $Q_X = F_X^{-1}$.
- In R, `q` functions: `qnorm`, `qexp`, `qbinom`.

## The inverse transform method

For **continuous** $X$ and $U \sim U(0,1)$:
$$Q_X(U) \sim X$$

The recipe:

1. Find the CDF $F(x)$.
2. Set $u = F(x)$ and solve for $x$. That is $Q(u)$.
3. Return $Q$ applied to `runif(n)`.

Exponential: $u = 1 - e^{-\lambda x}$ gives $Q(u) = -\log(1 - u)/\lambda$, so `-log(runif(n)) / rate`.

**Corollary** (probability integral transform): $F_X(X) \sim U(0,1)$.

## Checking a sampler: QQ plots

A quantile function **uniquely determines** a distribution.

- Plot sample quantiles against the target's quantiles, $Q(u)$ for many $u$.
- **On the 45 degree line**: the sampler matches the target.
- **Off the line**: it does not.
- `geom_qq(distribution = q_ours)` takes **any** quantile function, including one you wrote.

That last point checks the sampler against **your math**, not against R.

## Discrete variables: $Q$ is a step function

The continuous proof fails: there may be no $x$ with $F(x) = u$. The method still works.

$$Q(u) = x \quad \text{where} \quad F(x - 1) < u \le F(x)$$

```{r fig.width=6, fig.height=1.6}
d <- 1:9
Fd <- log10(d + 1)
u <- 0.5
ggplot() +
  geom_segment(aes(x = d, xend = d + 1, y = Fd, yend = Fd)) +
  geom_point(aes(x = d, y = Fd), size = 1.5) +
  geom_hline(yintercept = u, linetype = "dashed") +
  annotate("point", x = 3, y = Fd[3], size = 3, shape = 1) +
  scale_x_continuous(breaks = 1:9, limits = c(1, 10)) +
  labs(x = "Leading digit d", y = "F(d)") +
  theme_bw(base_size = 9)
```

Benford with $u = 0.5$: $F(2) = 0.477 < 0.5 \le 0.602 = F(3)$, so $d = 3$.

## Two ways to compute a discrete $Q$

**Walk the PMF.** Works for any PMF you can compute.

1. Draw $u$, set $v = 0$.
2. For $x_1 < x_2 < \ldots$: add $p(x_i)$ to $v$. Stop at the **first** $x_i$ with $v \ge u$.

**Closed form.** When $F$ has one, solve $F(x - 1) < u \le F(x)$ directly.

- Benford: $F(d) = \log_{10}(d + 1)$, so $Q(u) = \lceil 10^u - 1 \rceil$.
- In R: `ceiling(10^runif(n) - 1)`.
