---
title: "The Inverse Transform"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
institute: "Computational Methods in Statistics and Data Science (DATASCI 406)"
date: "Module 2: Generate"
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: false
    includes:
      in_header: ../preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%")
options(width = 60, digits = 4)
set.seed(20260929)
library(tidyverse)
```

# Review and Overview of Module 2

## Module 1 Review

In Module 1, we used **pre-existing** sampling functions to

- Estimate integrals, expectations, and probabilities
- Find more efficient Monte Carlo estimators (**lower variance**)
- Compare estimators for parameters on **bias**, **variance**, **mean squared error**.
- Construct hypothesis tests using **null and alternative distributions**, **compute $p$-values**, **invert** into a confidence interval
- Evaluate confidence intervals for **coverage** and **expected width**

## Module 2 Overview

In this module we **create our own sampling functions**

- Turn uniform draws into **any** distribution whose quantile function we can write
- **Transform** and **mix** variables we can already draw
- Sample a density we can only **evaluate**, by **rejecting** or **reweighting**
- **Factor** a joint distribution, draw one coordinate at a time
- Build a **Markov chain** whose long run behavior is the distribution we want

## Why generate random numbers?

Our data are already random. Why manufacture more?

- Monte Carlo **integration**, **estimation**, **hypothesis tests** (module 1)
- **Resampling**: bootstrap, permutation tests, cross validation (module 3)
- Non-deterministic **optimization** (module 4)
- **Random assignment** in trials, **random sampling** from a population
- **Simulating** a process to see what it produces
- **Noise** for privacy, **cryptography**

Each needs a sampler for a particular distribution.

# Pseudorandom Numbers

## Random and Pseudorandom Numbers

When we used `runif`, we claimed this generated **random variables uniformly distributed**.

An event is **random** if we cannot predict it, knowing everything we could know.

$$\underbrace{f(x) = a + bx}_{\text{deterministic}} \qquad
\underbrace{f(x) = a + bx + \epsilon, \; \epsilon \sim N(0,1)}_{\text{random}}$$

**True** randomness needs a physical source: atmospheric noise, thermal noise, radioactive decay.

Physical sources are slow, and they **cannot be reproduced**.

**Pseudorandom** numbers are deterministic, but close enough that no test we run tells the difference.

## Where would you get true randomness?

```{r eval=FALSE}
library(random)                 # random.org, atmospheric noise
randomNumbers(10, 0, 255)

runif(10, 0, 255)               # pseudorandom
```


## The entropy pool, in base R

macOS and Linux expose the OS entropy pool as a file. **Windows does not.**

```{r echo=FALSE, comment=NA}
cat(readLines("01-entropy-pool.txt"), sep = "\n")
```

Each $V$ is 16 random bits, so $V/2^{16}$ is our uniform. Code in `01-entropy-pool.R`.

## Comparison

| Source | Per call | Reproducible? |
|---|---:|---|
| random.org, over the network | 1.7 s | no |
| `/dev/urandom`, the OS entropy pool | 0.16 ms | no |
| `runif` | 0.005 ms | **yes** |

Speed matters, but one big feature is **reproducibility.** We can get the same random sequence by setting a "seed" value.

## Uniform bits

Computers represent numbers as groups of **bits**: values that are 0 or 1.

Generating a **(pseudorandom) number** is equivalent to producing random bits.

A uniform distribution on the integers $0$ to $2^{n} - 1$ needs $n$ IID bits with

$$P(B_i = 1) = P(B_i = 0) = 0.5$$

Let $V$ be the number encoded by those $n$ bits, then

$$U = \frac{V}{2^n} \sim \text{Uniform}(0, 1)$$

## Pseudorandom Number Generators (PRNGs)

Computers are **deterministic** so how do we get "random numbers?"

A **PRNG** takes a **seed** and returns a number and a **new seed**:

```{r eval=FALSE}
prng <- function(seed) {
  # do something
  c(random_number, new_seed)
}
```

Nothing here is random, but the sequence looks **as if** it were $U(0,1)$.

The seed is what makes it reproducible: `set.seed()` picks where you start.

## A very simple generator

A **linear congruential generator**:

```{r}
lcg <- function(seed) {
  a <- 5; c <- 1; m <- 8
  v <- (a * seed + c) %% m
  c(v/m, v) # random number, next seed in sequence
}
```

## LCG sequence

```{r}

rngs <- numeric(16)
seed <- 7
for (i in 1:16) {
  rs <- lcg(seed)
  rngs[i] <- rs[1]
  seed <- rs[2]
}
rngs
```

## Evaluating PRNGs 

This was not a great choice of a PRNG (but it was fast!)

- The sequence **repeats after 8 values**
- Each value fixes the next, so the **period** cannot exceed the modulus $m$
- After one cycle the draws are no longer plausibly independent

How would we know a generator is good?

- We **cannot prove** it. No finite test certifies randomness.
- We **can reject** violations, with the hypothesis tests like module 1 built
- Packages `TestU01` and `diehard` run hundreds of them

## PRNGs in R

R implements several; see `?Random`.

The default **Mersenne-Twister** has period $2^{19937} - 1$ (the LCG had $2^3$).

```{r}
set.seed(3949392) # fixes the starting point
runif(3)
```

From here we assume `runif()` works, and ask what we can build from it.

# From Uniform to Anything

## Distribution and quantile functions

The **(cumulative) distribution function** of a random variable is

$$F_X(t) = \Pr(X \le t).$$

It is **non-decreasing** and **right continuous**, with $0 \le F_X(t) \le 1$.

That lets us define the **quantile function**

$$Q_X(p) = \inf\left\{x : F_X(x) \ge p\right\}, \quad p \in [0,1].$$

When $F_X$ is continuous and strictly increasing, this is just $F_X^{-1}$.

## Example: uniform on $(0, \theta)$

For $X \sim U(0, \theta)$ with $f(x) = 1/\theta$ on $(0, \theta)$:

$$F(t) = \int_0^t \frac{1}{\theta} \, dx = \frac{t}{\theta}.$$

$X$ is continuous, so invert directly:

$$u = \frac{t}{\theta} \quad \Rightarrow \quad Q(u) = \theta u.$$

**Notice**: When we sampled `50 * runif(n)` to get $U(0, 50)$ that is the same as $Q(U)$ where $U \sim U(0,1)$. We'll prove the general case of this shortly.



## The inverse transform method

Let $X$ be a **continuous** random variable with quantile function $Q_X$.

For $U \sim \text{Uniform}(0,1)$,

$$Y = Q_X(U)$$

**has the same distribution as $X$.**

One uniform draw, one function evaluation, one sample from $X$.

## Proof

Since $X$ is continuous, $\inf\{x : F(x) \ge u\} = \inf\{x : F(x) = u\}$, so $F(\inf\{x : F(x) = u\}) = u$.

And $F_U(u) = u$.

\[
\begin{aligned}
\Pr(Q(U) \le x) &= \Pr(\inf\{x : F(x) \ge U\} \le x) & \text{(definition)} \\
&= \Pr(\inf\{x : F(x) = U\} \le x) & \text{(continuity)} \\
&= \Pr(F(\inf\{x : F(x) = U\}) \le F(x)) & \text{($F$ non-decreasing)} \\
&= \Pr(U \le F(x)) & \text{(first fact)} \\
&= F(x) = \Pr(X \le x) & \text{(second fact)}
\end{aligned}
\]

## Corollary: the probability integral transform

Run the argument backwards. For continuous $X$,

$$F_X(X) \sim U(0,1).$$

Feeding a random variable into **its own** CDF gives a uniform.

The inverse transform takes uniforms to $X$; this takes $X$ back to uniforms.

We use it to **test** samplers, and again in module 3.

## Example: the exponential

$$f(x) = \lambda e^{-\lambda x}, \quad x \ge 0, \quad \lambda > 0$$

$$F(x) = \int_0^x \lambda e^{-\lambda t} \, dt = 1 - e^{-\lambda x}$$

Solve $u = 1 - e^{-\lambda x}$ for $x$:

$$Q(u) = -\frac{\log(1 - u)}{\lambda}$$

If $U \sim U(0,1)$ then $1 - U \sim U(0,1)$, so $-\log(U)/\lambda$ works too.

## Implementing it

```{r}
rexp_ours <- function(n, rate = 1) {
  -log(runif(n)) / rate
}

x <- rexp_ours(100)
```


## Checking your work: QQ plots

Quantile functions **uniquely determine** a distribution.

$F_X$ and $F_Y$ may live on different scales, but $Q_X$ and $Q_Y$ both map $(0,1) \to \mathbb{R}$.

A **QQ plot** draws $(Q_X(u), Q_Y(u))$ for many $u \in (0,1)$.

If the distributions agree, the points land on the 45 degree line.

Gives visual test of correctness.

## QQ plot: right distribution

```{r fig.height=2, fig.width=6}
ggplot(data.frame(sample = x), aes(sample = sample)) +
  geom_qq(distribution = qexp) +
  geom_abline(intercept = 0, slope = 1)
```

## QQ plot: wrong distribution

```{r fig.height=2, fig.width=6}
ggplot(data.frame(sample = x), aes(sample = sample)) +
  geom_qq(distribution = qnorm) +
  geom_abline(intercept = 0, slope = 1)
```

## When R has no `q` function

`geom_qq` takes **any** quantile function, including one we wrote ourselves.

```{r fig.height=2, fig.width=6}
q_ours <- function(p, rate = 1) -log(1 - p) / rate

ggplot(data.frame(sample = x), aes(sample = sample)) +
  geom_qq(distribution = q_ours) +
  geom_abline(intercept = 0, slope = 1)
```

This checks the sampler against **our math**, not against R.


# Discrete Random Variables

## Why the proof breaks

The continuous proof used continuity of $F$ to write

$$Q(u) = \inf\{x : F(x) \ge u\} = \inf\{x : F(x) = u\}.$$

There had to be **some** point with $F(x) = u$.

For discrete $X$ there need not be: if $X \sim \text{Bin}(1, 0.5)$, no $x$ has $F(x) = 0.75$.

$Q(u)$ is still well defined. It is a **step function**.

We need a different argument and a different algorithm.

## $Q$ as a step function

For any discrete $X$,

$$Q(u) = \min\{x : F(x) \ge u\} = \min\left\{x : \sum_{i \le x} p(i) \ge u\right\}$$

So if $Q(u) = x$, then $\sum_{i \le x} p(i) \ge u$ and $\sum_{i \le x-1} p(i) < u$.

$$Q(u) = x \quad \text{such that} \quad F(x - 1) < u \le F(x).$$

The uniform lands in one of the **jumps** of $F$, and that jump names the value.

## The algorithm

$Q$ is a step function, so instead of building, tabulate probabilities until you pass $u$:

1. Generate $U \sim U(0,1)$, set $v = 0$
2. For $x_1 < x_2 < \ldots$:
    a. $v \leftarrow v + p(x_i)$
    b. if $v \ge u$, return $x_i$

Works for **any** discrete distribution with a computable PMF.


## Benford's Law, revisited

Recall, Benford's Law describes the leading digit $D$ for random variables that are products of independent components:

$$\Pr(D = d) = \log_{10}\left(\frac{d+1}{d}\right), \quad d = 1, \ldots, 9.$$

```{r echo=FALSE}
d <- 1:9
pd <- log10((d + 1) / d)
tab <- rbind("P(D = d)" = round(pd, 3), "F(d)" = round(cumsum(pd), 3))
colnames(tab) <- d
tab
```

$F(d)$ climbs to 1. A uniform draw lands in exactly one of those steps.


## Closed form CDF for Benford

While we often have to resort to $F(x) = \sum_{v \le x} P(X = v)$ for discrete CDFs, sometimes they have **closed form expressions**.

For example, Benford:
\[
\begin{aligned}
\Pr(D \le d) = \sum_{i=1}^d \log_{10}\left(\frac{i+1}{i}\right)
= \log_{10}\left(\frac{2 \times 3 \times \cdots \times (d+1)}{1 \times 2 \times \cdots \times d}\right)
= \log_{10}(d + 1)
\end{aligned}
\]

## The quantile function of $D$

This means we also get a closed form quantile function. Apply $F(d-1) < u \le F(d)$ with $F(d) = \log_{10}(d+1)$:

\[
\begin{aligned}
Q(u) &= d \text{ s.t. } \log_{10}(d) < u \le \log_{10}(d+1) \\
&= d \text{ s.t. } d < 10^{u} \le d + 1 \\
&= d \text{ s.t. } d - 1 < 10^{u} - 1 \le d \\
&= \left\lceil 10^u - 1 \right\rceil
\end{aligned}
\]

As before, draw $U \sim U(0,1)$ and apply $Q(U)$.

## Sampling from $D$

```{r}
rbenford <- function(n) ceiling(10^runif(n) - 1)

k <- 10000
rbind(analytic  = log10((2:10) / (1:9)),
      empirical = table(rbenford(k)) / k)
```

Compare `10^runif(n)` to $x = 10^m \cdot 10^f$ from the module 1 exercise.


## Summary

- A PRNG gives **uniform** draws. Everything else is built on them.
- $Q_X(p) = \inf\{x : F_X(x) \ge p\}$ is defined for **every** distribution
- **Continuous**: $Q_X(U) \sim X$. One uniform, one function call.
- **Corollary**: $F_X(X) \sim U(0,1)$, the probability integral transform
- **Discrete**: $Q$ is a step function. Walk the PMF, or find $Q$ in closed form.
- **Check every sampler with a QQ plot**

Next: what to do when we cannot write $Q$ down.
