---
title: "Rejection and Accept-Reject Sampling"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
institute: "Computational Methods in Statistics and Data Science (DATASCI 406)"
date: "Module 2: Generate"
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: false
    includes:
      in_header: ../preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%")
options(width = 60, digits = 4)
set.seed(20261006)
library(tidyverse)
```

# Rejection Sampling

## Rejection Sampling

<!-- TODO: opening slide placing day 3 in the module table: we have a density
(possibly only up to a constant) and an envelope, but no quantile function. -->

The idea of **rejection sampling** can be used when our **target RV** can be expressed as a subset of another **candidate RV**.

Examples:

- Conditioning on the variable itself, $X \sim \text{Poisson}(\lambda)$, $Y = X \mid X \text{ is odd}$
- Truncated distributions, e.g. $X \sim N(0, 1)$, $Y = X \mid a < X < b$
- Uniform points on the unit circle: $U_1, U_2 \sim U(-1, 1)$ (independent), $(V_1, V_2) = (U_1, U_2) \mid U_1^2 + U_2^2 \le 1$

Procedure: generate from candidate, only keep results that meet criteria.

## Example: Truncated Normal Distribution

A **truncated distribution** takes a given distribution and **limits the support**.

For example, $X \sim N(0,1)$ and $Y = X \mid 0 \le X \le 1$. We can sample by drawing from $X$ and only keeping those that fall in $(0, 1)$.

```{r}
x <- rnorm(10000)
y <- keep(x, 0 < x & x < 1)
length(y) # number of samples kept
mean(y) # use the samples to estimate E(Y)
```

## The truncated normal CDF

To check the sampler we need the distribution of $Y$. For $X \sim N(0,1)$, $Y = X \mid a \le X \le b$ and $a \le y \le b$,
$$F_Y(y) = P(X \le y \mid a \le X \le b) = \frac{P(a \le X \le y)}{P(a \le X \le b)} = \frac{\Phi(y) - \Phi(a)}{\Phi(b) - \Phi(a)}$$
where $\Phi$ is the standard normal CDF (`pnorm`).

The denominator is the probability that a draw is kept:

```{r}
pnorm(1) - pnorm(0)
```

So we throw away about two thirds of the draws from $X$.

## Truncated normal quantiles

Set $F_Y(y) = p$ and solve for $y$:
$$\Phi(y) = \Phi(a) + p\,[\Phi(b) - \Phi(a)] \Rightarrow Q_Y(p) = \Phi^{-1}\big(\Phi(a) + p\,[\Phi(b) - \Phi(a)]\big)$$

The quantiles of $Y$ come from the quantiles of the full normal (`qnorm`), evaluated at $p$ rescaled into $[\Phi(a), \Phi(b)]$.

```{r}
qtnorm <- function(p, a, b) {
  qnorm(pnorm(a) + p * (pnorm(b) - pnorm(a)))
}
```

## Checking the rejection sampler

```{r fig.height=2.5, fig.width=6}
ggplot(data.frame(sample = y), aes(sample = sample)) +
  geom_qq(distribution = qtnorm, dparams = list(a = 0, b = 1)) +
  geom_abline(intercept = 0, slope = 1)
```

## Rejection sampling summary

- The target is a candidate conditioned on an event: $Y = X \mid X \in A$.
- Draw from $X$ and keep the draws that land in $A$.
- The fraction kept is $P(X \in A)$, which is $\Phi(b) - \Phi(a)$ for the truncated normal. When $A$ is rare, most draws are wasted.
- Check the kept draws with a QQ plot against $Q_Y$.

# Accept-Reject Method

## Computing $\pi$

Consider the following method for approximating the value of $\pi$:

- Initialize $a = 0$
- Generate $U_1$ and $U_2$ (independent) from $U(0,1)$
- If $\sqrt{U_1^2 + U_2^2} \le 1$ then $a = a + 1$
- Repeat $k$ times.
- Approximate $\pi$ as
      $$\pi = \frac{4a}{k}$$

## Calculating in R

```{r}
approx_pi <- function(k) {
  u1 <- runif(k)
  u2 <- runif(k)
  sqs <- sqrt(u1^2 + u2^2)
  return(4 * sum(sqs <= 1) / k)
}
```

## Approximation vs. $k$

```{r echo=FALSE, fig.height=3.4, fig.width=6}
xs <- seq(100, 100000, length.out = 50)
ggplot(data.frame(k = xs, estimate = sapply(xs, approx_pi)), aes(k, estimate)) +
  geom_line() +
  geom_hline(yintercept = pi, lty = 2, color = "red") +
  labs(y = "Approx Pi")
```

## Why does it work?

Recall: the circle centered at $(0, 0)$ and with radius 1, has area $\pi$.

We "throw darts" at the upper quadrant of the circle and see how many hit the circle.

This proportion should be about $\pi/4$

##

```{r echo=FALSE, fig.height=3, fig.width=3, out.width="50%", fig.align="center"}
u1 <- runif(100)
u2 <- runif(100)
inside <- u1^2 + u2^2 <= 1
ggplot(data.frame(u1, u2, inside)) +
  geom_point(aes(u1, u2, color = inside, shape = inside)) +
  geom_function(fun = function(x) sqrt(1 - x^2), xlim = c(0, 1), n = 501, linewidth = 1) +
  scale_color_manual(values = c("red", "blue")) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  theme(legend.position = "none")
```

## Another interpretation of Approximating $\pi$

Notice that the following is a valid **probability density function**:
$$f(x) = \frac{4}{\pi} \sqrt{1 - x^2}, 0 \le x \le 1$$

This curve is just a **scaled version of the unit circle**.

##

```{r echo=FALSE, fig.height=3.5, fig.width=4, out.width="50%", fig.align="center"}
u1 <- runif(100)
u2 <- runif(100)
inside <- u1^2 + u2^2 <= 1
ggplot(data.frame(u1, u2, inside)) +
  geom_point(aes(u1, 4 / pi * u2, color = inside, shape = inside)) +
  geom_function(fun = function(x) 4 / pi * sqrt(1 - x^2), xlim = c(0, 1), n = 501, linewidth = 1) +
  scale_color_manual(values = c("red", "blue")) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 4 / pi)) +
  labs(x = NULL, y = NULL) +
  theme(legend.position = "none")
```

##

```{r echo=FALSE, fig.height=3.5, fig.width=4, out.width="50%", fig.align="center"}
k <- 10000
u1 <- runif(k)
u2 <- runif(k)
inside <- u1^2 + u2^2 <= 1
ggplot(data.frame(u1, u2, inside)) +
  geom_point(aes(u1, 4 / pi * u2, color = inside, shape = inside), size = 0.5) +
  geom_function(fun = function(x) 4 / pi * sqrt(1 - x^2), xlim = c(0, 1), n = 501, linewidth = 1) +
  scale_color_manual(values = c("red", "blue")) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 4 / pi)) +
  labs(x = NULL, y = NULL) +
  theme(legend.position = "none")
```

##

```{r echo=FALSE, fig.height=3.8, fig.width=6}
ggplot(data.frame(x = u2[inside]), aes(x = x )) + geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "blue", alpha = 0.5) + stat_function(fun = function(x) { 4/pi*sqrt(1 - x^2)}, lty = 2, col = "red", lwd = 2)
```

## Interpretation

- We started by drawing $U_1$ and $U_2$, and then keeping $U_2$ if $U_1^2 + U_2^2 \le 1$.
- This is equivalent to picking $U_2$, and keeping $U_1$ if
      $$\frac{4}{\pi} U_2 \le \frac{4}{\pi} \sqrt{1 - U_1^2} = f(U_1)$$
      or written another way:
      $$ U_2 \le \frac{\pi}{4} \frac{f(U_1)}{g(U_1)}, \quad g(x) = 1 \text{ (pdf of $U_1$)}$$
- This process generated samples from $X$ (the variable with density $f(x)$)!

## Accept-Reject in General

Suppose we want to sample from density (or mass) function:
$$f(x)$$

We know $f(x)$, but can't easily sample from it directly (e.g., Normal distribution).

But what if we had **another density** $g(y)$ such that
$$c \times \frac{g(x)}{f(x)} \ge 1$$
for some $c > 0$ and for all $x$ such that $f(x) > 0$.

In other words we need $c g(x)$ to lie above $f(x)$ for any $x$ where $f(x)$ is positive.

## Accept-reject for $f(x)$

For the example we just did the **target density** was
$$f(x) = \frac{4}{\pi} \sqrt{1 - x^2}, 0 \le x \le 1$$

The **candidate density** was the uniform distribution
$g(x) = 1, 0 \le x \le 1$

The **constant** was
$$c = \frac{4}{\pi}$$

This ensures that:
$$c g(x) = \frac{4}{\pi} \ge \frac{4}{\pi}\sqrt{1 - x^2}  = f(x)$$
(NB: $c$ is not unique. We could have picked any $c \ge 4/\pi$.)

## Example: Uniform and Beta

Suppose we want to sample from the Beta(2,2) distribution which is:
$$f(x) = 6x(1-x), \quad 0 \le x \le 1$$
but we only have the standard uniform $g(x) = 1$.

Observe that $f(x)$ achieves its max at $x = 0.5$, the maximum value is $f(0.5) = 6/4$. Therefore we have
$$\frac{6}{4} \frac{g(x)}{f(x)} \ge 1$$
for any point in $[0,1]$.

##

```{r echo=FALSE, fig.width=6, fig.height=4}
ggplot() +
  geom_function(fun = function(x) 6 * x * (1 - x), xlim = c(0, 1), n = 501) +
  geom_hline(yintercept = 1, lty = 2, color = "blue") +
  geom_hline(yintercept = 6/4, lty = 2, color = "red") +
  ylim(0, 1.5) +
  labs(x = "x", y = "density")
```

## Accept-Reject Algorithm

- Draw a $Y$ from $g(y)$
- Draw $U \sim U(0,1)$
- If $$ U < \frac{f(Y)}{c g(Y)}$$
      then **accept** $X = Y$
- Otherwise, **reject** $Y$ as a candidate and repeat.

## Example: Beta/Uniform

Say we generate $Y = 0.3$.

- $f(0.3) = 1.26$
- $g(0.3) = 1$
- We will accept if
      $$U < \frac{1.26}{(6/4) \times 1} = 0.84$$

##

```{r echo=FALSE, fig.width=6, fig.height=3}
ggplot() +
  geom_function(fun = function(x) 6 * x * (1 - x), xlim = c(0, 1), n = 501) +
  geom_hline(yintercept = 6/4, lty = 2, color = "blue") +
  annotate("segment", x = 0.3, xend = 0.3, y = 0, yend = 1.26, linewidth = 1.5, color = "green") +
  annotate("segment", x = 0.3, xend = 0.3, y = 1.26, yend = 6/4, linewidth = 1.5, color = "red") +
  ylim(0, 1.5) +
  labs(x = NULL, y = NULL)
```

## R implementation

```{r}
k <- 10000
ys <- runif(k)
fys <- 6 * ys * (1 - ys)
gys <- 1
ratios <- fys / (gys * (6/4))
```

```{r}
us <- runif(k)
accept <- us < ratios
accepted <- ys[accept]
rejected <- ys[!accept] ; mean(!accept)
```

##

```{r echo=FALSE, fig.width=6, fig.height=3}
ggplot(data.frame(y = ys, u = us, accept)) +
  geom_point(aes(y, u * 6/4, color = accept, shape = accept)) +
  geom_function(fun = function(x) 6 * x * (1 - x), xlim = c(0, 1), n = 501, linewidth = 1) +
  geom_hline(yintercept = 6/4, lty = 2, color = "blue") +
  scale_color_manual(values = c("red", "blue")) +
  ylim(0, 1.5) +
  labs(x = NULL, y = NULL)
```

## Checking the accepted draws

```{r fig.height=2.5, fig.width=6}
ggplot(data.frame(sample = ys[accept]), aes(sample = sample)) +
  geom_qq(distribution = qbeta, dparams = list(shape1 = 2, shape2 = 2)) +
  geom_abline(intercept = 0, slope = 1)
```

## Proving the general case

When proving an algorithm generates $X$ we need to show same cumulative distribution function (recall, we did this for inversion method, transformations, etc.).

Let $V$ be the random variable produced by the AR algorithm, we need to show that
$$P(V \le x) = P(X \le x)$$
for any $x$ in the support of $X$.

## What kind of variable is $V$?

We know the distribution of $Y$ and $U$, but what can we say about $V$?

Notice that $V$ is set to $Y$, when $U \le f(Y) / (c g(Y))$, a conditional distribution:
$$V = Y \, \left| \, U \le \frac{f(Y)}{c \, g(Y)} \right.$$

For events $A$ and $B$, recall the definition of conditional probability:
$$P(A | B) = \frac{P(A, B)}{P(B)}$$

## Proof

\[
P(V \le x) = P\left(Y \le x  \, \left| \, U \le \frac{f(Y)}{c \, g(Y)} \right.\right)
    = \frac{P\left(Y \le x, U \le \frac{f(Y)}{c \, g(Y)} \right)}{P\left(U \le \frac{f(Y)}{c \, g(Y)}\right)}
\]

We'll take the numerator and denominator separately.

In both cases, think of writing $P(\ldots) = E(I(\ldots))$, e.g.:
$$P\left(Y \le x, U \le \frac{f(Y)}{c \, g(Y)} \right) = \E{I\left(Y \le x, U \le \frac{f(Y)}{c \, g(Y)}\right)}$$

## Numerator

Notice (a) by independence, the joint density is $1 \times g(y)$ and (b) $I(A, B) = I(A) I(B)$.
\[
\begin{aligned}
P\left(Y \le x, U \le \frac{f(Y)}{c \, g(Y)} \right) &= \int_{-\infty}^{\infty} \int_0^1 I(y \le x) I(u \le f(y) / (c g(y))) g(y) \, du \, dy  \\
&= \int_{-\infty}^{\infty} I(y \le x) g(y) \left[\int_0^1  I(u \le f(y) / (c g(y))) \, du \right] \, dy  \\
&= \int_{-\infty}^{x} g(y) \left[\int_0^{f(y) / (c g(y))} 1 \, du \right] \, dy  \\
&= \int_{-\infty}^{x} g(y) \frac{f(y)}{ c g(y)} \, dy   \\
&=  \frac{1}{c} \int_{-\infty}^{x}  f(y) \, dy  = \frac{1}{c} P(X \le x)
\end{aligned}
\]

## Denominator

$$P\left(U \le \frac{f(Y)}{c \, g(Y)} \right) = \E{I\left(U \le \frac{f(Y)}{c \, g(Y)}\right)}$$
By similar logic,
\[
\begin{aligned}
P\left(U \le \frac{f(Y)}{c \, g(Y)} \right) &= \int_{-\infty}^{\infty} g(y) \left[\int_0^{f(y) / (c g(y))} 1 \, du \right] \, dy  \\
&= \int_{-\infty}^{\infty} g(y) \frac{f(y)}{c g(y)} \, dy   \\
&= \int_{-\infty}^{\infty} \frac{1}{c} f(y) \, dy  = \frac{1}{c}
\end{aligned}
\]

Taking the ratio, the $(1/c)$ cancels, so $P(V \le x) = P(X \le x)$.

## Example: Truncated Normal

We've used the **rejection sampling algorithm** for the Truncated Normal before:
$$Y = Z \mid 0 \le Z \le 1, Z \sim N(0,1)$$

To use AR, we need the **density function**. We found the CDF earlier:
$$F_Y(y) = \frac{\Phi(y) - \Phi(0)}{\Phi(1) - \Phi(0)}, \quad 0 \le y \le 1$$

The PDF is its derivative. With $\phi$ the standard normal PDF, for $0 \le y \le 1$,
$$f(y) = \frac{\phi(y)}{\Phi(1) - \Phi(0)}$$

## Implementing in R

```{r}
(scaling_const <- pnorm(1) - pnorm(0))
```

Then we compute the density function as:

```{r}
truncated <- function(x) {
  ifelse(x >= 0 & x <= 1,
         dnorm(x) / scaling_const,
         0)
}
```

## Picking a candidate density: Uniform

```{r echo=FALSE, fig.width=6, fig.height=3}
envelope <- data.frame(x = c(0, 0, 1, 1), y = c(0, truncated(0), truncated(0), 0))
ggplot() +
  geom_function(fun = truncated, xlim = c(-0.5, 1.5), n = 501, linewidth = 1) +
  geom_path(data = envelope, aes(x, y), lty = 2, linewidth = 1, color = "blue") +
  labs(x = "x", y = "density")
```

## Implementing

```{r}
ys <- runif(1000)
const <- truncated(0)
ratios <- truncated(ys) / (const * 1) # g(y) = 1
```

```{r}
us <- runif(1000)
accept_uniform <- us < ratios
accepted_uniform <- ys[accept_uniform]
rejected_uniform <- ys[!accept_uniform]
```

Estimating $\E{X}$:

```{r}
mean(accepted_uniform)
```

##

```{r echo=FALSE, fig.width=6, fig.height=3}
ggplot() +
  geom_point(data = data.frame(y = ys, u = us, accept = accept_uniform),
             aes(y, u * truncated(0), color = accept, shape = accept)) +
  geom_function(fun = truncated, xlim = c(-0.5, 1.5), n = 501, linewidth = 1) +
  geom_path(data = envelope, aes(x, y), lty = 2, linewidth = 1, color = "blue") +
  scale_color_manual(values = c("red", "blue")) +
  labs(x = NULL, y = NULL)
```

We rejected $`r 100 * round(1 - mean(accept_uniform), 3)`\%$ of the candidate draws. Can we do better?

## Fitting a closer candidate distribution

Consider the density:
$$g(y) = \frac{2}{3}(2 - y), \quad 0 \le y \le 1$$
with quantile function
$$Q_y(p) = 2 - \sqrt{4 - 3 p}$$

Need to find a $c$ such that
$$c \times \frac{g(x)}{f(x)} \ge 1 \Rightarrow c \ge \frac{f(x)}{g(x)}, 0 \le x \le 1$$

##

Notice that the **ratio $f(x)/g(x)$ is increasing**:

```{r echo=FALSE, fig.height=2, fig.width=6}
g <- function(y) { (2/3) * (2 - y)}
ggplot() +
  geom_function(fun = function(x) dnorm(x) / g(x), xlim = c(0, 1)) +
  labs(x = "x", y = "f(x)/g(x)")
```

Then the maximum ratio is at $x = 1$.
$$c = f(1)/g(1) = `r round(truncated(1)/g(1), 3)` \Rightarrow \frac{c g(x)}{f(x)} \ge 1, 0 \le x \le 1$$

##

```{r}
g <- function(y) { (2/3) * (2 - y)}
qg <- function(p) { 2 - sqrt(4 - 3 * p)}
ys <- qg(runif(1000))
const <- truncated(1) / g(1)
ratios <- truncated(ys) / (const * g(ys))
```

```{r}
us <- runif(1000)
accept_g <- us < ratios
accepted_g <- ys[accept_g]
rejected_g <- ys[!accept_g]
```

##

```{r echo=FALSE, fig.width=6, fig.height=3}
envelope_g <- data.frame(x = c(0, 0, 1, 1), y = c(0, const * g(0), const * g(1), 0))
ggplot() +
  geom_point(data = data.frame(y = ys, u = us, accept = accept_g),
             aes(y, u * const * g(y), color = accept, shape = accept)) +
  geom_function(fun = truncated, xlim = c(-0.5, 1.5), n = 501, linewidth = 1) +
  geom_path(data = envelope_g, aes(x, y), lty = 2, linewidth = 1, color = "blue") +
  scale_color_manual(values = c("red", "blue")) +
  labs(x = NULL, y = NULL)
```

Now we reject only $`r 100 * round(1 - mean(accept_g),3)`\%$ of the candidate draws.

## Comparing Efficiency

Recall: **more efficient** means lower variance. Both envelopes give exact draws from $f$, and from $m$ candidates, accept-reject keeps about $N = m/c$, so the mean of the kept draws has
$$\Var{\bar X} = \frac{\Var{X}}{N} \approx \frac{c \, \Var{X}}{m}.$$

So for a fixed $m$, the ratio of the variances is the ratio of the constants:

```{r}
c_uniform <- truncated(0)
c_g <- truncated(1) / g(1)
(rel_eff <- c_uniform / c_g)
```

The uniform envelope's estimate has `r round(100 * (rel_eff - 1))`\% more variance than the estimate using $g$.

## A bimodal density

Suppose we have a density of the form:
$$f(x) = 0.25 \phi(x + 2) + 0.75 \phi(x - 1), -\infty < x < \infty$$
where $\phi$ is the standard Normal PDF.

```{r}
f <- function(x) { 0.25 * dnorm(x + 2) + 0.75 * dnorm(x - 1)}
```

A density like this is called **bimodal** because it contains two local maxima.

## Density of $f$

```{r echo=FALSE, fig.height=3, fig.width=6}
ggplot() + geom_function(fun = f, xlim = c(-5, 4), n = 501) + labs(x = "x", y = "f(x)")
```

## What candidate distribution?

Our initial thought might be to use $\phi(x)$ (standard Normal) as the candidate.

The only problem with that is that
$$\lim_{x \rightarrow \pm\infty} f(x)/\phi(x) = \infty$$
(i.e., no $c$ exists that uniformly bounds $f$).

We saw that the Cauchy distribution has **fat tails**, perhaps that would be useful?
$$g(x) = \frac{1}{\pi (1 + x^2)}$$

## Ratio of $f(x)/\phi(x)$

```{r echo=FALSE, fig.height=3, fig.width=6}
ggplot() + geom_function(fun = function(x) f(x) / dnorm(x), xlim = c(-10, 18), n = 501) +
  labs(x = "x", y = "f(x) / phi(x)")
```

## Ratio of $f(x)/g(x)$

```{r echo=FALSE, fig.height=3, fig.width=6}
ggplot() + geom_function(fun = function(x) f(x) / dcauchy(x), xlim = c(-10, 18), n = 501) +
  labs(x = "x", y = "f(x) / g(x)")
```

## Lazy mode: find $c$ by evaluation

```{r}
h <- function(x) { f(x) / dcauchy(x)}
xs <- seq(-4, 4, length.out = 1000)
(const <- max(h(xs)))
```

## Plotting Distributions

```{r echo=FALSE, fig.height=3, fig.width=6}
ggplot() +
  geom_function(aes(linetype = "c * Cauchy"), fun = function(x) const * dcauchy(x), xlim = c(-5, 5), n = 501) +
  geom_function(aes(linetype = "f(x)"), fun = f, xlim = c(-5, 5), n = 501) +
  labs(x = "x", y = "density", linetype = NULL)
```

## Putting it together

```{r}
k <- 10000
ys <- rcauchy(k)
ratios <- f(ys) / (const * dcauchy(ys))
us <- runif(k)
accept <- us < ratios
x <- ys[accept] ; mean(accept)
```

##

```{r echo=FALSE, warning=FALSE, fig.height=3.8, fig.width=6}
df <- data.frame(y = ys, us = us, ratios, accept = accept, scaleu = const * dcauchy(ys) * us )
ggplot(df, aes(x = ys, y = scaleu, color = accept)) +
    xlim(-5,5) +
    geom_point() +
    stat_function(fun = f, lty = 2, col = "black") +
    stat_function(fun = ~ const * dcauchy(.x), lty = 1, col = "red") +
    labs(x = "Y", y = "c * g(Y) * U")
```

```{r echo=FALSE}
rm(ratios) # just to avoid bugs in some copy/pasted code later
```

## What if we got $c$ wrong?

We found $c = `r round(const, 3)`$. What if we were wrong? How would the distribution change? Too large, too small?

```{r}
wrong_consts <- c(2 *  const,
                  const,
                  0.75 * const,
                  0.25 * const,
                  0.1 * const,
                  0)
```

## Functions to handle AR

For numerical reasons, we'll rewrite $U \le f(Y) / (c  g(Y))$ as
$$c \, g(Y) \, U \le f(Y)$$

```{r}
accept_reject <- function(n, f, g, rg, const) {
    ys <- rg(n)
    us <- runif(n)
    accept <- us * const * g(ys) <= f(ys)
    data.frame(y = ys, accept = accept)
}
```

## Using the bad constants

```{r}
bimodal <- function(const) {
    accept_reject(1000, f, dcauchy, rcauchy, const)
}
wrong <- map_dfr(wrong_consts, .id = "constant", bimodal)
wrong$constant <- wrong_consts[as.numeric(wrong$constant)] # get the original labels back
```

##

```{r echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=6}
accept_rates <- group_by(wrong, constant) %>% summarize(rate = mean(accept)) %>% mutate(label = factor(paste0(round(constant, 2), " (", round(rate, 2), ")")))

wrong_accepted <- wrong[wrong$accept, ] %>% left_join(accept_rates, by = "constant")
ggplot(wrong_accepted, aes(x = y)) + geom_density(fill = "blue", alpha = 0.25, lty = 0) +
    stat_function(fun = f, lty = 2, lwd = 1, color = "red") +
    stat_function(fun = dcauchy, lty = 1, lwd = 1, color = "#008800") +
    xlim(-5, 5) +
    facet_wrap(~ label)
```

<!-- TODO: failure case from the README, an envelope whose tail is too light.
Use a normal envelope for a Cauchy target: no finite c bounds f/g, so any
chosen c accepts draws whose tails are too thin. Show it as a sampling failure
(QQ plot of accepted draws against qcauchy), not as an IS variance problem. -->

## Accept-Reject Summary

- For a density $f$, use a density $g$ for the sampling.
- Draw uniform values to decide to accept or reject draws from $g$.
- Need to figure out the scaling constant $c$ to make $c g(y) / f(y) \ge 1$
- Generally, we accept about $100 \times (1/c)\%$ of the candidates, so making $c$ small is useful.
- Picking $c$ too large is ok, too small leads to draws from $g$.

<!-- TODO: accept-reject for an unnormalized f. AR also works with f* in
place of f, provided c bounds f*/g; the unknown constant is absorbed into c.
Importance resampling, the other method for a density known up to a constant,
is in deck 5. -->

<!-- TODO: deck summary slide covering rejection and accept-reject, plus a
"Next" line pointing to day 4. -->
