---
title: "Turn-in 2.3: Project Part C"
author: "Your Name, Your Uniqname"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
set.seed(20260929)
library(tidyverse)
```

Worth 1 point for completion.

Fill in the blanks in each code chunk and write your answers after each
**Answer:**. Knit to HTML and upload the `.html` file to Canvas.

A chunk with an unfilled blank shows an error when you knit. That is expected
until you fill it in.

These come from the project:

```{r definitions}
rlaplace <- function(n) {
  u <- runif(n)
  ifelse(u < 1/2, log(2 * u), -log(2 - 2 * u))
}
fstar <- function(x) sin(pi * x^2)
```

## Question 1 (C.1a)

Find the smallest $c$ with $c \ge \phi(x)/g(x)$ for all $x$. The ratio
$\phi(x)/g(x)$ is symmetric in $x$, so take $x \ge 0$, where
$\phi(x) = e^{-x^2/2}/\sqrt{2\pi}$ and $g(x) = e^{-x}/2$.

(a) Take logs. $\log\dfrac{\phi(x)}{g(x)}$ is a constant plus a term in $x$.
What are the two?

**Answer:**

(b) Take the derivative with respect to $x$.

**Answer:**

(c) Where is the derivative zero?

**Answer:**

(d) Plug that $x$ into the ratio. What is $c = \phi(x)/g(x)$?

**Answer:**

## Question 2 (C.1b)

Fill in the blanks, then make a QQ plot of `z` against the normal
(`geom_qq()` uses the normal by default).

```{r c1b}
c_lap <- ________                  # in R, e is exp(1)
y <- rlaplace(10000)
keep <- runif(10000) <= dnorm(y) / (c_lap * ________)   # c times g(y)
z <- y[keep]

# your QQ plot of z here

mean(keep)                          # acceptance rate
1 / c_lap
t.test(as.numeric(keep))$conf.int   # 95% interval for the acceptance rate
```

**Checkpoint:** put your QQ plot of the accepted draws on screen.

Does the 95% interval contain $1/c$?

**Answer:**

## Question 3 (C.2)

**We will do C.2(a) and the related part of C.2(c). Importance resampling has been moved to day 5.**

**Note.** Accept-reject works even though $f^*$ is not a proper density.
Write $f = k\,f^*$, where $k$ is the unknown normalizing constant, and pick
$c$ with $c\,g(x) \ge f^*(x)$ for all $x$. In the proof, the probability of
accepting a draw at or below $x$ is $\frac{1}{ck} F(x)$. The probability of
accepting at all is $\frac{1}{ck}$. Their ratio is $F(x)$, because the $k$s
cancel. The accepted draws come from $f$. Only the acceptance rate changes:
it is $1/(ck)$ instead of $1/c$.

Fill in the blank.

```{r c2}
y1 <- runif(10000)
ar <- y1[runif(10000) <= ________]        # (a) accept-reject, c = 1
exact <- integrate(function(x) x * fstar(x), 0, 1)$value /
  integrate(fstar, 0, 1)$value             # (c)
mean(ar)
exact
```

## Question 4 (C.1)

Suppose you had used $c = 2$ in C.1. Would the accepted draws still be
$N(0, 1)$? What would happen to the acceptance rate?

**Answer:**

