---
title: "Multivariate Distributions and Processes"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
institute: "Computational Methods in Statistics and Data Science (DATASCI 406)"
date: "Module 2: Generate"
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: false
    includes:
      in_header: ../preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%")
options(width = 60, digits = 4)
set.seed(20261008)
library(tidyverse)
```

# Joint Distributions

## Factoring a joint distribution

The definition of a conditional density, $f(x_2 \mid x_1) = f(x_1, x_2) / f(x_1)$, rearranges to
$$f(x_1, x_2) = f(x_1) \, f(x_2 \mid x_1).$$

Applying it repeatedly **factors** any joint density into a product of conditionals:
$$f(x_1, \ldots, x_d) = f(x_1) \, f(x_2 \mid x_1) \, f(x_3 \mid x_1, x_2) \cdots f(x_d \mid x_1, \ldots, x_{d-1})$$

Any ordering of the coordinates works.

## Sampling one coordinate at a time

The factorization gives a sampler:

1. Draw $X_1$ from $f(x_1)$.
2. Draw $X_2$ from $f(x_2 \mid X_1)$.
3. Continue until $X_d$ is drawn from $f(x_d \mid X_1, \ldots, X_{d-1})$.

Each step is a **univariate** problem, so the methods from days 1--3 apply.

Day 2's mixtures were this with $d = 2$: draw the component or the parameter, then draw $X$ given it.

## Example: $f(x, y) = x + y$ on the unit square

The marginal of $X$, its CDF, and its quantile function:
$$f(x) = \int_0^1 (x + y)\,dy = x + \frac{1}{2}, \quad F(x) = \frac{x^2 + x}{2}, \quad Q_X(u) = \frac{-1 + \sqrt{1 + 8u}}{2}$$

The conditional of $Y$ given $X = x$:
$$f(y \mid x) = \frac{x + y}{x + \frac{1}{2}}, \quad F(y \mid x) = \frac{xy + y^2/2}{x + \frac{1}{2}}, \quad Q(u \mid x) = -x + \sqrt{x^2 + u(2x + 1)}$$

Each quantile function comes from solving a quadratic for the root in $[0, 1]$.

## In R

```{r}
qx <- function(p) (-1 + sqrt(1 + 8 * p)) / 2
qy_given_x <- function(p, x) -x + sqrt(x^2 + p * (2 * x + 1))

n <- 10000
x <- qx(runif(n))
y <- qy_given_x(runif(n), x) # y[i] is drawn given x[i]
```

## Checking the marginal of $Y$

By symmetry $Y$ has the same marginal as $X$, although we never drew $Y$ from its marginal.

```{r fig.height=2.2, fig.width=6}
ggplot(data.frame(sample = y), aes(sample = sample)) +
  geom_qq(distribution = qx) +
  geom_abline(intercept = 0, slope = 1)
```

## Checking the dependence

A QQ plot checks one coordinate at a time, so it would also pass if we drew $Y$ independently of $X$. The correlation checks the dependence.

From $f$: $\E{XY} = 1/3$, $\E{X} = \E{Y} = 7/12$, and $\Var{X} = 11/144$, so
$$\text{Cor}(X, Y) = \frac{1/3 - (7/12)^2}{11/144} = -\frac{1}{11}.$$

```{r}
c(exact = -1/11, factored = cor(x, y),
  independent = cor(x, qx(runif(n))))
```



# Multivariate Normal Distributions

## Linear Algebra Refresher

This section will use **vectors** and **matrices** in describing **multivariate distributions**.

A **vector** is a collection of values indexed $1, 2, \ldots, k$. E.g. $X_3$ is third element of $\bm{X}$.

A **matrix** is an $n \times m$ collection of values. $X_{23}$ indicates the entry at the second row, third column of $\bm{X}$.

We can think of a matrix as **$m$ vectors of length $n$**. Or a vector as a **$n \times 1$ matrix**.

The **transpose of a matrix** swaps the rows and columns. E.g., $X_{ij} = (\bm{X}^T)_{ji}$.


## Matrix multiplication

We can **multiply matrices with conformable dimensions** such that if $\bm{A}$ is $n \times m$ and $\bm{B}$ is $m \times k$, we **can multiply $\bm{A}\bm{B}$** but **not $\bm{B}\bm{A}$**.

Let $\bm{C} = \bm{A}\bm{B}$, then
$$C_{ij} = \sum_{k = 1}^m  A_{ik} B_{kj} \quad \text{sum of pairwise products $i$ row of $\bm{A}$ with $j$ column of $\bm{B}$}$$
The **identity matrix** is a special $n \times n$ matrix such that
$$\bm{I}_n \bm{A} = \bm{A} \bm{I}_m = \bm{A}$$

A **matrix inverse** (if it exists) is such that for $\bm{A}$, $n \times n$:
$$\bm{A}\bm{A}^{-1} = \bm{I}_n$$


## Multivariate Normal (MVN) Distributions

A special case of multivariate distributions with a very simple dependence structure is the multivariate distribution is the multivariate Normal distribution:
$$\bm{X} \sim N_d(\bm{\mu}, \bm{\Sigma})$$


- $\bm{\mu}$ is a vector of length $d$ such that $\E{\bm{X}} = \bm{\mu}$ (the **mean**) -- $\E{X_i} = \mu_i, i = 1, \ldots, d$
- $\bm{\Sigma}$ is the $d \times d$ **variance-covariance matrix** such that $\Cov{X_i}{X_j} = \Sigma_{ij} = \Sigma_{ji}$
- When $\bm{\Sigma}$ is positive definite (usually), the PDF is:
$$f(\bm{x}) = |2\pi \bm{\Sigma}|^{-1/2} \exp\left\{-\frac{1}{2} (\bm{x} - \bm{\mu})^T \bm{\Sigma}^{-1} (\bm{x} - \bm{\mu})\right\}$$

## Motivation: many correlated tests

We test $m = 10$ outcomes measured on the same subjects. Under the null, the test statistics are $\bm{Z} \sim N_m(\bm{0}, \bm{\Sigma})$, with correlation induced by testing same subjects.

**Bonferroni** rejects $H_{0i}$ when $|Z_i| > \Phi^{-1}(1 - \alpha/2m)$. Its family-wise error rate
$$\text{FWER} = P\left(\max_i |Z_i| > \Phi^{-1}(1 - \alpha/2m)\right)$$
is at most $\alpha$ for any $\bm{\Sigma}$. For independent tests it is $1 - (1 - \alpha/m)^m$.

How far below $\alpha$ is it when the tests are correlated? In general the FWER is a 10-dimensional normal integral. Monte Carlo can estimate it if we can draw from $N_m(\bm{0}, \bm{\Sigma})$.


## Key intuition: what is the square root of $\Sigma$?

We know we can generate $X \sim N(\mu, \sigma^2)$ as
$$X = \sigma Z + \mu, \quad Z \sim N(0, 1)$$

For univariate $Z$, to get $X$ we multiply $Z$ by the square root of $\sigma^2$ (and add $\mu$).

Think about the square root as the function that finds $a$ such that $a \times a = b$:
$$\sqrt{b} = a : a \times a = b$$
We will do the same thing by finding the "square root" of the matrix $\bm{\Sigma}$: two matrices that multiply to $\bm{\Sigma} = \bm{C}\bm{C}^T$.

## Generating MVN

Observe that we can generate the vector $\bm{Z}$ as a series of univariate, independent standard Normals:
$$\bm{Z} \sim N_d(\bm{0}, \bm{I}_d)$$
where $I_d$ is the identity matrix:
\[
\begin{pmatrix}
    1 & 0 & \cdots & 0 \\
    0 & 1 & \cdots & \vdots \\
    \vdots & 0 & \cdots & 1
\end{pmatrix}
\]

If we can factor $\bm{\Sigma} = \bm{C}\bm{C}^T$, then by the linearity of multivariate Normals:
$$\bm{C} \bm{Z} + \bm{\mu} \sim N_d(\bm{\mu}, \bm{C} I_d \bm{C}^T)  =  N_d(\bm{\mu}, \bm{\Sigma})$$

## Pre-Proof

Some definitions. Let $C_{ij}$ be the $i$ row and $j$ column of $C$. If we define $\bm{\Sigma} = \bm{C}\bm{C}^T$,
\[
\begin{array}{lllr}
\Var{X_i} &= \Sigma_{ii}   &= \sum_{k = 1}^d C_{ik}^2 & \text{(sum of squared entries in $i$th row)} \\
\Cov{X_i}{X_j} &= \Sigma_{ij}   &= \sum_{k = 1}^d C_{ik} C_{jk} & \text{(sum of pairwise products for $i$ and $j$)}
\end{array}
\]

Some facts:

- $Z_1, Z_2, \ldots Z_d \iid N(0, 1)$
- $Z_i^2 \sim \chi_1^2$, $\E{Z_i^2} = 1$
- For any RV $Y$, $\Var{c Y} = c^2 \Var{Y}$
- For independent RVs, $\Var{Y_1 + Y_2} = \Var{Y_1} + \Var{Y_2}$ and $\E{Y_1 Y_2} = \E{Y_1} \E{Y_2}$

## Proof: Marginal distribution

Claim: $\bm{X} = \bm{C} \bm{Z} \sim N_d(\bm{0}_d, \bm{\Sigma})$. First, show marginal distribution:
\[
\begin{aligned}
    X_1 &= C_{11} Z_1 + C_{12} Z_2 + \ldots C_{1d} Z_d \sim N(0, \sum_{k=1}^d C_{1k}^2) = N(0, \Sigma_{11}) \\
    X_2 &= C_{21} Z_1 + C_{22} Z_2 + \ldots C_{2d} Z_d \sim N(0, \sum_{k=1}^d C_{2k}^2) = N(0, \Sigma_{22})
\end{aligned}
\]

## Proof: Covariance

We won't prove joint Normality, but we will show correct covariance:
\[
\begin{aligned}
    \Cov{X_1}{X_2} &= \E{X_1 X_2} - \E{X_1} \E{ X_2}  \\
    &= \E{\left(\sum_{k=1}^d C_{1k} Z_k\right) \left(\sum_{l=1}^d C_{2l} Z_l\right)}  \\
    &= \E{ \sum_{k = 1}^d \sum_{l = 1}^d C_{1k} C_{2l} Z_k Z_l }  \\
    &= \sum_{k = 1}^d \sum_{l = 1}^d C_{1k} C_{2l} \E{Z_k Z_l}  \\
    &= \sum_{k = 1}^d C_{1k} C_{2k} \E{Z_k^2}  + \sum_{k \ne l} C_{1k} C_{2l} \E{Z_k Z_l}
    =  \sum_{k = 1}^d C_{1k} C_{2k} = \Sigma_{12}
\end{aligned}
\]


## Factoring $\bm{\Sigma}$

Any positive definite $\bm{\Sigma}$, which includes most covariance matrices, has a **Cholesky factorization**
$$\bm{\Sigma} = \bm{C}\bm{C}^T, \quad \bm{C} \text{ lower triangular.}$$

In R, `chol(Sigma)` returns the upper triangular $\bm{C}^T$, so use `C <- t(chol(Sigma))`.

## The Cholesky decomposition as factoring

For two coordinates with unit variances and correlation $\rho$:
$$\bm{C} = \begin{pmatrix} 1 & 0 \\ \rho & \sqrt{1 - \rho^2} \end{pmatrix}, \qquad \bm{X} = \bm{C}\bm{Z}: \quad X_1 = Z_1, \quad X_2 = \rho X_1 + \sqrt{1 - \rho^2}\, Z_2$$

- $X_1$ is drawn from its marginal, $N(0, 1)$.
- $X_2$ is drawn from its conditional given $X_1$, $N(\rho X_1, 1 - \rho^2)$.

Each row of a lower triangular $\bm{C}$ adds one new $Z_i$, so in general $X_i$ is drawn given $X_1, \ldots, X_{i-1}$. In other words, using the factoring we started with.

## Back to the tests

```{r}
m <- 10
bonferroni <- qnorm(1 - 0.05 / (2 * m))

max_abs_z <- function(rho, B = 100000) {
  Sigma <- matrix(rho, m, m)
  diag(Sigma) <- 1
  C <- t(chol(Sigma))
  Z <- C %*% matrix(rnorm(m * B), m) # each column is a draw
  apply(abs(Z), 2, max)
}
```

## Bonferroni with correlated tests

```{r}
results <- tibble(rho = c(0, 0.3, 0.6, 0.9)) %>%
  mutate(maxes = map(rho, max_abs_z),
         fwer = map_dbl(maxes, ~ mean(.x > bonferroni)),
         cutoff = map_dbl(maxes, ~ quantile(.x, 0.95)))
select(results, -maxes)
```

## What the simulation shows

At $\rho = 0$ the exact FWER is $1 - 0.995^{10} = 0.0489$, which checks the sampler.

As $\rho$ grows, Bonferroni's FWER falls well below 5%, so it rejects less often than it could.

Rejecting when $|Z_i|$ exceeds `cutoff`, the 95th percentile of $\max_i |Z_i|$, gives a FWER of 5% (Bonferroni uses `r round(bonferroni, 2)`).

## Data: blue crabs

`MASS::crabs` has five size measurements, in mm, of 50 male and 50 female blue crabs: frontal lobe (FL), rear width (RW), carapace length (CL) and width (CW), and body depth (BD). Do the sexes differ? That is five tests.

```{r}
blue <- subset(MASS::crabs, sp == "B")
vars <- c("FL", "RW", "CL", "CW", "BD")
male <- as.matrix(blue[blue$sex == "M", vars])
female <- as.matrix(blue[blue$sex == "F", vars])
z <- (colMeans(male) - colMeans(female)) /
  sqrt(apply(male, 2, var) / 50 + apply(female, 2, var) / 50)
round(z, 2)
```

## The null distribution for the crabs

Under the null, $\bm{Z} \approx N_5(\bm{0}, \bm{R})$, where $\bm{R}$ is the correlation of the measurements within a sex. Every correlation is above 0.95.

```{r}
# center each sex at its own mean, then correlate
R <- cor(rbind(scale(male, scale = FALSE), scale(female, scale = FALSE)))
C <- t(chol(R))
maxes <- apply(abs(C %*% matrix(rnorm(5 * 100000), 5)), 2, max)
quantile(maxes, 0.95)
```

Bonferroni's cutoff for five tests is `r round(qnorm(1 - 0.05 / 10), 2)`.

## Adjusted $p$-values

Each method gives an adjusted $p$-value, and we reject when it is below 0.05:

- Bonferroni: $\min(1, m \, p_i)$
- Max-$|Z|$: $P(\max_j |Z_j| \ge |z_i|)$, estimated by the fraction of `maxes` at least $|z_i|$

```{r}
p <- 2 * pnorm(-abs(z))
max_z <- sapply(abs(z), function(t) mean(maxes >= t))
round(cbind(z, p, bonferroni = pmin(1, 5 * p), max_z), 3)
```

## What the crabs show

Body depth (BD) is significant using the max-$|Z|$ cutoff but not using Bonferroni.

Every max-$|Z|$ $p$-value is about a third of the Bonferroni one. With correlations near 1, the five measurements nearly measure one thing, so the max-$|Z|$ $p$-values are close to the unadjusted ones.

## MVN Summary

- General dependence is **difficult to categorize**.
- The Normal distribution equates **dependence to covariance/correlation** (defining characteristic of the distribution).
- Also serves a **common approximation** (via CLT)
- Extend the idea of $Z \sim N(0, 1) \Rightarrow X = \sigma Z + \mu \sim N(\mu, \sigma^2)$, to
      $$\bm{Z} \sim N_d(\bm{0}, \bm{I}_d) \Rightarrow \bm{C} \bm{Z} + \bm{\mu} \sim N_d(\bm{\mu}, \bm{C}\bm{C}^T)$$
- Need a concept the **square root of a matrix** (e.g., Cholesky factorization)

<!-- TODO: mixtures of multivariate normals, one sentence only. The project
works a full example (faithful, D.1-D.2), so the video should not. -->

<!-- TODO: uniform distribution on the d-sphere as Z / |Z| for
Z ~ N_d(0, I_d). -->

<!-- TODO: Wishart distribution, mention only. -->

<!-- TODO: failure case, dimension. Rejection sampling on the d-ball accepts
52% of draws at d = 3 and 1 in 40 million at d = 20. -->

# Stochastic Processes

## Stochastic Processes

A **stochastic process** is a collection of random variables indexed by time or space:
$$\{X(t) : t \in \mathcal{T}\}$$

The index set $\mathcal{T}$ can be discrete, such as $\{0, 1, 2, \ldots\}$, or continuous, such as $[0, \infty)$.

To simulate a process we draw $X(t)$ at finitely many times. That is a random vector, so we factor its joint distribution and draw one time at a time.

## Example: Random Walk

- Start from $X(0) = x_0$.
- For $t = 1, 2, \ldots$, set
      $$X(t) = X(t - 1) + S(t)$$
      where $P(S(t) = 1) = P(S(t) = -1) = 0.5$ and the steps $S(t)$ are independent.

Unrolling the recursion,
$$X(t) = x_0 + \sum_{i=1}^t S(i).$$

## The random walk is a factorization

Factor the walk's joint distribution over time:
$$f(x_1, \ldots, x_k \mid x_0) = f(x_1 \mid x_0) \, f(x_2 \mid x_1, x_0) \cdots f(x_k \mid x_{k-1}, \ldots, x_0)$$

Each factor depends only on the previous value:
$$f(x_t \mid x_{t-1}, \ldots, x_0) = f(x_t \mid x_{t-1}), \qquad X(t) \mid X(t - 1) = x \text{ is } x \pm 1 \text{ with probability } 1/2$$

A process with dependence limited previous steps is a **Markov chain**. 

## Implementation

```{r fig.height=2, fig.width=6}
s <- sample(c(1, -1), 100, replace = TRUE)
x <- c(0, cumsum(s))
ggplot(tibble(t = 0:100, x), aes(t, x)) + geom_step()
```

## Many walks

$X(t)$ is a sum of $t$ independent steps with variance 1, so $\Var{X(t)} = t$, and by the CLT $X(t)/\sqrt{t} \approx N(0, 1)$.

```{r fig.height=1.7, fig.width=6}
walk_end <- function(k) sum(sample(c(1, -1), k, replace = TRUE))
ends <- replicate(5000, walk_end(400))
ggplot(tibble(sample = ends / sqrt(400)), aes(sample = sample)) +
  geom_qq() + geom_abline(intercept = 0, slope = 1)
```

## Brownian motion

Shrink the steps: take a step every $\Delta t$, with variance $\Delta t$. As $\Delta t \to 0$ the walk converges to **Brownian motion** $W(t)$, $t \ge 0$:

- $W(0) = 0$
- $W(t) - W(s) \sim N(0, t - s)$ for $s < t$
- Increments over non-overlapping intervals are independent.

So $W(t) \sim N(0, t)$, and the paths are continuous but nowhere smooth.

## Simulating Brownian motion

We can only draw $W$ on a grid $0 = t_0 < t_1 < \cdots < t_n$. Factoring as for the walk,
$$W(t_i) \mid W(t_{i-1}) \sim N\big(W(t_{i-1}), \, t_i - t_{i-1}\big).$$

```{r}
brownian <- function(n, t_max = 1) {
  dt <- t_max / n
  tibble(t = seq(0, t_max, length.out = n + 1),
         w = c(0, cumsum(rnorm(n, sd = sqrt(dt)))))
}
```

## Five paths

```{r echo=FALSE, fig.height=3, fig.width=6}
paths <- map_dfr(1:5, ~ brownian(1000), .id = "path")
ggplot(paths, aes(t, w, color = path)) + geom_line() + theme(legend.position = "none")
```

## Checking the sampler: the maximum

The maximum $M = \max_{0 \le t \le 1} W(t)$ has the same distribution as $|Z|$ for $Z \sim N(0, 1)$ (the **reflection principle**), so
$$Q_M(p) = \Phi^{-1}\left(\frac{1 + p}{2}\right).$$

```{r}
q_max <- function(p) qnorm((1 + p) / 2)
max_w <- function(n) max(brownian(n)$w)
sim_max <- tibble(n = rep(c(20, 1000), each = 2000)) %>%
  mutate(m = map_dbl(n, max_w))
```

## A coarse grid misses the peaks

```{r fig.height=1.8, fig.width=6}
ggplot(sim_max, aes(sample = m, color = factor(n))) +
  geom_qq(distribution = q_max) +
  geom_abline(intercept = 0, slope = 1) + labs(color = "grid size")
```

With 20 grid points the maximum is too small, because the peaks fall between grid points.

## Brownian bridge

A **Brownian bridge** is Brownian motion on $[0, 1]$ conditioned to return to 0: $B(t) = W(t) \mid W(1) = 0$.

Any ordering of the coordinates works, so draw the endpoint first and the middle given the ends:
$$W(t) \mid W(1) = w_1 \sim N\big(t \, w_1, \; t(1 - t)\big)$$

Subtracting the straight line from 0 to $W(1)$ removes the endpoint, so $B(t) = W(t) - t \, W(1)$.

```{r}
bridge <- function(n) {
  mutate(brownian(n), b = w - t * last(w))
}
```

## Five bridges

```{r echo=FALSE, fig.height=3, fig.width=6}
bridges <- map_dfr(1:5, ~ bridge(1000), .id = "path")
ggplot(bridges, aes(t, b, color = path)) + geom_line() + theme(legend.position = "none")
```

<!-- Callback for later: when the Kolmogorov-Smirnov test is taught (the
permutation-test lectures in fa25), point back here. sqrt(n) (F_n - F)
behaves like a Brownian bridge evaluated at F(x), so the KS null
distribution is that of max_t |B(t)|. -->

## Checking the bridge

The maximum $\max_t B(t)$ has $P(\max_t B(t) > a) = e^{-2a^2}$, so
$$Q(p) = \sqrt{-\log(1 - p) / 2}.$$

```{r fig.height=1.35, fig.width=6}
q_bridge_max <- function(p) sqrt(-log(1 - p) / 2)
bridge_max <- replicate(2000, max(bridge(1000)$b))
ggplot(tibble(sample = bridge_max), aes(sample = sample)) +
  geom_qq(distribution = q_bridge_max) +
  geom_abline(intercept = 0, slope = 1)
```

## Poisson process

Events arrive at times $T_1 < T_2 < \cdots$, and $N(t)$ counts the arrivals up to time $t$.

- The gaps $T_k - T_{k-1}$ are independent $\text{Exp}(\lambda)$, drawn by inversion as on day 1.
- Each arrival time is a sum of gaps, $T_k = E_1 + \cdots + E_k$, as on day 2.
- Then $N(t) \sim \text{Poisson}(\lambda t)$.

Factoring again: $T_k \mid T_{k-1} = s$ is $s$ plus an $\text{Exp}(\lambda)$ gap.

## Implementation

```{r}
arrivals <- function(lambda, t_max) {
  times <- c()
  t <- rexp(1, lambda)
  while (t <= t_max) {
    times <- c(times, t)
    t <- t + rexp(1, lambda)
  }
  times
}
```

## One path of $N(t)$

```{r fig.height=2.2, fig.width=6}
ggplot(tibble(t = c(0, arrivals(2, 10))), aes(t, seq_along(t) - 1)) +
  geom_step() + labs(y = "N(t)")
```

## Checking the count

With $\lambda = 2$ and $t = 10$, $N(10) \sim \text{Poisson}(20)$.

```{r fig.height=2.2, fig.width=6}
counts <- replicate(5000, length(arrivals(2, 10)))
ggplot(tibble(sample = counts), aes(sample = sample)) +
  geom_qq(distribution = qpois, dparams = list(lambda = 20)) +
  geom_abline(intercept = 0, slope = 1)
```

## Some other stochastic processes

- **Martingale**: $\E{X(t) \mid X(t - 1), X(t - 2), \ldots} = X(t - 1)$. The symmetric random walk is one.
- **Spatial Poisson process**: in a region of area $A$, draw $N \sim \text{Poisson}(\lambda A)$ and place $N$ points uniformly at random.
- **Autoregressive processes**: $X(t) = \phi X(t - 1) + \epsilon(t)$, which reduces to the random walk at $\phi = 1$.

## Stochastic process summary

- A process is a collection of dependent random variables $\{X(t)\}$, indexed by time or space.
- We simulate one at a time, drawing each $X(t)$ given the earlier ones. This is the factorization from the start of the deck.
- The random walk, Brownian motion and the Poisson process depend on the past only through the most recent value.
- A continuous-time process has to be drawn on a grid, and a coarse grid can bias what we estimate.

