---
title: "Turn-in 2.4: Project Part D"
author: "Your Name, Your Uniqname"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
set.seed(20260929)
library(tidyverse)
```

\newcommand{\E}[1]{\text{E}\left(#1\right)}

Worth 1 point for completion.

Fill in the blanks in each code chunk and write your answers after each
**Answer:**. Knit to HTML and upload the `.html` file to Canvas.

A chunk with an unfilled blank shows an error when you knit. That is expected
until you fill it in.

The S&P 500 data from the project:

```{r spy}
load("../../data/spy.rda")
logp <- log(spy$price)
r <- diff(logp)
```

## Question 1 (D.1)

Fill in the blanks, then plot the data next to 272 points from `rfaithful`.

```{r d1}
short <- faithful$eruptions < 3
p <- mean(short)
mu1 <- colMeans(faithful[short, ]);  S1 <- cov(faithful[short, ])
mu2 <- __________________________;  S2 <- __________________________

rmvn <- function(n, mu, Sigma) {
  C <- chol(Sigma)
  Z <- matrix(rnorm(2 * n), nrow = 2)
  t(mu + ________ %*% Z)                # mu + C^T Z, one draw per column
}
rfaithful <- function(n) {
  n1 <- rbinom(1, n, p)                  # how many short eruptions
  x <- rbind(rmvn(n1, mu1, S1), rmvn(n - n1, ____, ____))
  colnames(x) <- c("eruptions", "waiting")
  x
}

# your plot of the data next to rfaithful(272) here

p
mu1["waiting"]                           # mean wait after a short eruption
```

**Checkpoint:** put the data next to 272 points from `rfaithful` on screen.

## Question 2 (D.2a)

Show that the mean of the mixture is $p\mu_1 + (1 - p)\mu_2$. Write a draw as
$X = WU + (1 - W)V$, with $W \sim \text{Bernoulli}(p)$ independent of
$U \sim N_2(\mu_1, \Sigma_1)$ and $V \sim N_2(\mu_2, \Sigma_2)$.

$$\E{X} = \E{WU} + \E{(1 - W)V} = \E{W}\,\E{U} + \text{(blank 1)} = \text{(blank 2)}$$

What goes in the two blanks?

**Answer:**

Which step uses independence?

**Answer:**

## Question 3 (D.2b)

Compare the variance of `waiting` in each.

```{r d2b}
cov(faithful)
cov(rfaithful(10000))
p * S1 + (1 - p) * S2
```

Is the covariance of a mixture the weighted average of the component
covariances? What does the weighted average leave out?

**Answer:**

## Question 4 (D.3a)

Fill in the blanks, then plot 20 simulated paths with the data.

```{r d3a}
mu <- mean(r); sigma <- sd(r)
rwalk <- function() logp[1] + cumsum(c(0, rnorm(length(r), ____, ____)))
matplot(replicate(20, rwalk()), type = "l", lty = 1, col = "gray")
lines(logp)
```

**Checkpoint:** put 20 simulated paths with the observed log price on screen.

## Question 5 (D.3b, c, d)

Fill in the blank. Both $p$-values use the same 1,000 simulated paths.

```{r d3bc}
drawdown <- function(L) min(L - cummax(L))
abs_cor <- function(L) { a <- abs(diff(L)); cor(a[-1], a[-length(a)]) }
walks <- replicate(1000, rwalk(), simplify = FALSE)
dd <- sapply(walks, drawdown)
ac <- sapply(walks, abs_cor)
drawdown(logp)
(1 + sum(dd <= drawdown(logp))) / 1001   # (b) larger drawdowns
abs_cor(logp)
(1 + sum(ac ____ abs_cor(logp))) / 1001  # (c) more clustering
```

(d) Which feature of the data does the random walk reproduce, and which does
it miss?

**Answer:**

