---
title: "Multivariate Distributions and Processes: Recap"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
institute: "Computational Methods in Statistics and Data Science (DATASCI 406)"
date: "Module 2: Generate, Session 4"
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: false
    includes:
      in_header: ../preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = "100%")
library(ggplot2)
set.seed(20261008)
```

## Factoring a joint distribution

Any joint density factors into a product of conditionals:
$$f(x_1, \ldots, x_d) = f(x_1) \, f(x_2 \mid x_1) \cdots f(x_d \mid x_1, \ldots, x_{d-1})$$

- **Sampler**: draw $X_1$, then $X_2$ given $X_1$, and so on. Each step is a univariate problem from days 1--3.
- Any ordering of the coordinates works.
- **Checking**: a QQ plot checks one coordinate at a time, and would also pass if the coordinates were drawn independently. Check the dependence too, e.g. with a correlation.

## Multivariate normal

If $\bm{Z} \sim N_d(\bm{0}, \bm{I}_d)$ and $\bm{\Sigma} = \bm{C}\bm{C}^T$, then
$$\bm{C}\bm{Z} + \bm{\mu} \sim N_d(\bm{\mu}, \bm{\Sigma}).$$

- The **Cholesky factorization** gives a lower triangular $\bm{C}$. In R, `chol(Sigma)` returns $\bm{C}^T$, so use `C <- t(chol(Sigma))`.
- Cholesky is factoring. With unit variances and correlation $\rho$:
$$X_1 = Z_1, \qquad X_2 = \rho X_1 + \sqrt{1 - \rho^2}\, Z_2,$$
so $X_2$ is drawn from its conditional $N(\rho X_1, 1 - \rho^2)$.

## Application: correlated tests

$m$ test statistics on the same subjects, $\bm{Z} \sim N_m(\bm{0}, \bm{\Sigma})$ under the null.

- **Bonferroni** rejects when $|Z_i| > \Phi^{-1}(1 - \alpha/2m)$. Its FWER is at most $\alpha$, and far below it when the tests are correlated.
- **Max-$|Z|$**: reject when $|Z_i|$ exceeds the 95th percentile of $\max_i |Z_i|$, estimated by simulating $\bm{Z}$. Its FWER is 5%.

Blue crabs, five size measurements, correlations above 0.95:

| | Bonferroni | max-$\lvert Z \rvert$ |
|---|:-:|:-:|
| cutoff | 2.58 | 2.10 |
| body depth, $z = 2.57$ | not significant | significant |

## Stochastic processes

A process $\{X(t)\}$ is simulated at finitely many times, one time at a time, by factoring.

- **Random walk**: $X(t) = X(t - 1) \pm 1$. It depends on the past only through $X(t - 1)$, which makes it a **Markov chain**.
- **Brownian motion**: on a grid, $W(t_i) \mid W(t_{i-1}) \sim N\big(W(t_{i-1}), t_i - t_{i-1}\big)$.
- **Brownian bridge**: $B(t) = W(t) - t\,W(1)$, Brownian motion conditioned on $W(1) = 0$.
- **Poisson process**: $\text{Exp}(\lambda)$ gaps by inversion (day 1), arrival times as sums of gaps (day 2), and $N(t) \sim \text{Poisson}(\lambda t)$.

## Failure case: a coarse grid

The maximum of $W$ on $[0, 1]$ has the distribution of $|Z|$ (the reflection principle).

```{r fig.width=6, fig.height=2}
max_w <- function(n) max(cumsum(rnorm(n, sd = sqrt(1 / n))), 0)
sims <- data.frame(n = rep(c(20, 1000), each = 2000))
sims$m <- sapply(sims$n, max_w)
ggplot(sims, aes(sample = m, color = factor(n))) +
  geom_qq(distribution = function(p) qnorm((1 + p) / 2), size = 0.5) +
  geom_abline(intercept = 0, slope = 1) +
  labs(x = "quantiles of |Z|", y = "simulated maximum", color = "grid size") +
  theme_bw(base_size = 9)
```

Every grid value is an exact draw of $W(t_i)$, but the peaks fall between grid points, so the maximum on a coarse grid is too small.
