---
title: "Importance Resampling and MCMC: Recap"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
institute: "Computational Methods in Statistics and Data Science (DATASCI 406)"
date: "Module 2: Generate, Session 5"
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: false
    includes:
      in_header: ../preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = "100%")
library(ggplot2)
set.seed(20261013)
```

## Known up to a constant

$$f(x) = c \, f^*(x), \qquad c = \frac{1}{\int f^*(x) \, dx}$$

- $f^*$ is easy to evaluate, but the integral may not be.
- Today's two samplers use only $f^*$.

| | needs an envelope $g$? | draws |
|---|---|---|
| **Importance resampling** | yes | resampled from $g$'s candidates |
| **Markov chain Monte Carlo** | no | a dependent sequence |

## Importance resampling

1. Draw $Y_1, \ldots, Y_m$ from $g$.
1. Weight each by $f^*(Y_i)/g(Y_i)$, and normalize the weights to sum to 1.
1. Resample with `sample(y, replace = TRUE, prob = w)`.

- $c$ **cancels** when the weights are normalized.
- Nothing is rejected. The cost is a small bias that shrinks as $m$ grows.
- $g$ must cover $f$, with tails at least as heavy. Otherwise a few
  candidates carry almost all the weight.

Example: $Z \mid Z \ge 4.5$ with candidates $4.5 + \text{Exp}(1)$.

## Markov chains and burn in

A **Markov chain** draws $X(t)$ given only $X(t - 1)$. If $f$ is
**stationary** and the chain can reach all of $f$'s support, averages along
the chain converge to $E_f(h(X))$.

```{r fig.width=6, fig.height=1.8}
chain <- numeric(100); chain[1] <- 10
for (t in 2:100) chain[t] <- 0.9 * chain[t - 1] + sqrt(1 - 0.81) * rnorm(1)
ggplot(data.frame(t = 1:100, x = chain), aes(t, x)) + geom_line() +
  geom_hline(yintercept = c(-2, 2), lty = 2, color = "red") +
  labs(y = "X(t)", title = "X(t) = 0.9 X(t-1) + sqrt(0.19) Z(t), started at 10") +
  theme_bw(base_size = 9)
```

The early draws are not from $f$. Dropping them is **burn in**.

## Metropolis-Hastings

Accept-reject, with a candidate that depends on where the chain is.

1. Propose $X^* \sim g(\cdot \mid X(t - 1))$.
1. Move to $X^*$ if
   $$U \le \frac{f^*(X^*)}{f^*(X(t - 1))} \, \frac{g(X(t - 1) \mid X^*)}{g(X^* \mid X(t - 1))},$$
   and otherwise stay: $X(t) = X(t - 1)$.

- $f$ appears only in a ratio, so $c$ cancels.
- A move to lower density is sometimes accepted. That is what makes it a sampler and not an optimizer.
- Check with a **trace plot**, and with a QQ plot when $Q$ is known.

## Gibbs sampling

The **full conditional** of $x_i$ is its distribution given all the other
coordinates. As a function of $x_i$ it is proportional to $f$, so drop the
factors without $x_i$.

- Step through the coordinates, drawing each from its full conditional
  given the **current** values of the rest.
- A special case of Metropolis-Hastings in which every candidate is accepted.
- Needs full conditionals we can draw from directly, and needs burn in like
  any chain.

Bivariate normal, correlation 0.8: $X_1 \mid X_2 = t \sim N(0.8\,t,\ 0.36)$.

## Bayes and the Challenger

$$\pi(\theta \mid x) \propto f(x \mid \theta)\, p(\theta)$$
The posterior is known up to a constant, so all three samplers apply.

Logistic model for O-ring distress against launch temperature, 23 flights:

- Metropolis-Hastings gives $P(\text{distress at } 31^\circ\text{F})$ between
  0.88 and 1.00 (95%), and a grid agrees.
- Every flight was at least $22^\circ$F warmer, so the answer leans on the
  model's form outside the data.

**Failure case.** Importance resampling with the prior as the envelope:
the prior is far wider than the posterior, so a few candidates carry nearly
all the weight. The effective sample size is about 540 out of 100,000.
