---
title: "Monte Carlo Integration: Session Exercises"
subtitle: "Module 1, Day 1 --- DATASCI 406, Computational Methods in Statistics and Data Science"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 3.5,
                      collapse = TRUE, comment = "#>")
library(tidyverse)
set.seed(40601)
```

\newcommand{\E}[1]{\text{E}\left(#1\right)}
\newcommand{\Var}[1]{\text{Var}\left(#1\right)}
\newcommand{\ind}{\mathbf{1}}

Exercise 1 builds your toolbox of samplers. Every exercise after it uses
the recipe from the video:

1. Write the integral as $\E{h(X)}$ for a random variable $X$ you can sample.
2. Sample $X_1, \ldots, X_n$ and average $h(X_i)$.
3. Report a confidence interval, not just the point estimate.

Each integration exercise has a known true value, so you can check your
work. Your estimates will differ from the targets --- that is the point.
Your confidence interval should usually cover the target.

---

# 1. A tour of R's distributions

Let's investigate the distributions available in R.
`?Distributions` lists every distribution R ships with; each has the
`r`/`d`/`p`/`q` functions from the video. 

## Exercise

**(a)** Here is a start on a catalog of draws, in the long format that
`ggplot` likes:

```{r}
n <- 1000
draws <- list(
  norm   = rnorm(n),
  unif   = runif(n),
  exp    = rexp(n),
  cauchy = rcauchy(n)
) |> enframe() |> unnest(value)

draws |> group_by(name) |> summarize(mean(value))
```

Extend the list with **at least six more** distributions from the
`?Distributions` page, mixing continuous and discrete. Most need
parameters --- check the help page (e.g. `?rgamma`) and pick any sensible
values; skip `rmultinom` (it returns a matrix). Then plot the whole
catalog as histograms in one faceted figure
(`facet_wrap(~ name, scales = "free")`).

The **Cauchy** panel will look broken --- one bar and a huge empty axis.
It is not a bug. Investigate:

- Re-run your `summarize()` of the sample means a few times. Every other
  distribution's mean barely moves between runs. What does the Cauchy's do?
- Look at `sort(abs(rcauchy(n)), decreasing = TRUE)[1:5]` a few times.
  Where is the histogram's x-axis coming from?

**(b)** Pick **two** distributions from your catalog and explore their
parameters, plotting draws across a grid of parameter values. Here is the
pattern, applied to the Normal --- `expand_grid()` builds every
combination, and `map2()` draws a sample for each row:

```{r fig.height = 3}
expand_grid(mean = c(-5, 0, 5), sd = c(0.5, 2)) |>
  mutate(value = map2(mean, sd, ~ rnorm(2000, mean = .x, sd = .y))) |>
  unnest(value) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_grid(sd ~ mean, labeller = label_both)
```

Adapt it to two distributions of your choice (not the Normal). For at
least one of them, find parameter settings that make it match a
distribution you already know.

---

# 2. An integral on [0, 1]

For an integral over $[0,1]$, $\int_0^1 g(x)\,dx = \E{g(U)}$, where $U$ is uniform on 
$[0, 1]$. We can estimate the value of this integral by taking a sample of IID uniform draws
and computing the sample mean of $g(U_i)$.

## Exercise

Estimate
$$\theta = \int_0^1 \sin(\pi x) \, dx$$
using 10,000 uniform draws. Give a 95\% confidence interval using `t.test`.

The true value is $2/\pi \approx 0.6366$. Does your interval cover it?

---

# 3. Same integral, two RVs

The general recipe works with **any** density $f$ that is positive on the
domain: $\int g(x)\,dx = \E{g(X)/f(X)}$ for $X \sim f$. Different choices
of $f$ give different estimators of the **same** integral. Are they equally
good?

## Exercise

Estimate
$$\theta = \int_0^\infty x^2 e^{-x} \, dx$$
(true value $\Gamma(3) = 2! = 2$) twice, with 10,000 draws each:

**(a)** with $f = $ Exp(1) draws (`rexp` / `dexp`), and

**(b)** with $f = $ Gamma(shape 2, rate 1) draws (`rgamma` / `dgamma`) ---
the Gamma family from exercise 1.

Before coding, simplify $h(x) = g(x)/f(x)$ by hand in each case --- both
simplify pleasantly.

**(c)** Report a 95\% confidence interval for each. Both estimate the same
$\theta$ from the same number of draws --- which interval is narrower, and
by roughly what factor?

**(d)** Plot the **ECDFs** of the 10,000 $h$ values from (a) and the
10,000 from (b) on one set of axes (`stat_ecdf()`). What do the two curves
have in common, and how does their difference explain (c)?

**(e)** Repeat with $f = $ Gamma(shape 3, rate 1). Something
strange happens --- run it, then explain.

---

# 4. Probabilities are expectations of indicators

Recall an indicator function is a function that has the value 1 if a statement
is true and 0 if the statement is false. They are the mathematical equivalent of
`if` statements and conditionals in R. An important result is that for event $A$,

$$P(A) = E(\mathbf{1}(A))$$

In other words, we can estimate probabilities in the same way we estimate expectations.

## Exercise

For $Z \sim N(0, 1)$, estimate
$$\theta = P(1 \le Z \le 2)$$
using 100,000 draws.

1. Write the indicator directly in R (no `ifelse` needed).
2. Give a 99\% confidence interval. Which is the right tool here, `t.test`
   or `binom.test`, and why?
3. Check against the exact value from `pnorm`.

---

# 5. Joint distributions

Nothing in the recipe is one-dimensional. A double integral is an
expectation over a random **pair**: sample both coordinates, apply $h$,
average. When the coordinates are independent, sampling the pair is just
two calls to an `r` function.

## Exercise

**(a)** For $X, Y$ independent $U(0, 1)$, estimate the double integral
$$\theta = \int_0^1 \int_0^1 |x - y| \, dx \, dy = \E{|X - Y|}$$
with 100,000 pairs and a 95\% confidence interval. The true value is $1/3$.

**(b)** For $Z_1, Z_2$ independent $N(0, 1)$, estimate
$$P(Z_1 < Z_2 + 1)$$
with a 99\% confidence interval, using the right interval tool for the
job. Then check it exactly: what is the distribution of $Z_1 - Z_2$?

---

# 6. When the recipe breaks

The video stated conditions: $\E{h(X)} < \infty$ **and** $\E{h(X)^2} <
\infty$. This exercise is about what happens when the second one fails.

## Exercise

Estimate
$$\theta = \int_0^\infty \frac{1}{1 + x^2} \, dx$$
using Exp(1) draws, so $h(x) = e^{x} / (1 + x^2)$. The true value is
$\pi/2 \approx 1.5708$.

1. Run the estimator with $n = 100{,}000$ **five separate times**
   (`replicate` helps). Do the answers agree with each other? With the
   truth?
2. The estimator is unbiased --- we proved that in the video. So what is
   going wrong? Look at $\E{h(X)^2}$.
3. Would increasing $n$ fix this?

---
