---
title: "Monte Carlo Integration: Session Exercises"
subtitle: "Module 1, Day 1 --- DATASCI 406, Computational Methods in Statistics and Data Science"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 3.5,
                      collapse = TRUE, comment = "#>")
library(tidyverse)
set.seed(40601)
```

\newcommand{\E}[1]{\text{E}\left(#1\right)}
\newcommand{\Var}[1]{\text{Var}\left(#1\right)}
\newcommand{\ind}{\mathbf{1}}

Exercise 1 builds your toolbox of samplers. Every exercise after it uses
the recipe from the video:

1. Write the integral as $\E{h(X)}$ for a random variable $X$ you can sample.
2. Sample $X_1, \ldots, X_n$ and average $h(X_i)$.
3. Report a confidence interval, not just the point estimate.

Each integration exercise has a known true value, so you can check your
work. Your estimates will differ from the targets --- that is the point.
Your confidence interval should usually cover the target.

---

# 1. A tour of R's distributions

Let's investigate the distributions available in R.
`?Distributions` lists every distribution R ships with; each has the
`r`/`d`/`p`/`q` functions from the video. 

## Exercise

**(a)** Here is a start on a catalog of draws, in the long format that
`ggplot` likes:

```{r}
n <- 1000
draws <- list(
  norm   = rnorm(n),
  unif   = runif(n),
  exp    = rexp(n),
  cauchy = rcauchy(n)
) |> enframe() |> unnest(value)

draws |> group_by(name) |> summarize(mean(value))
```

Extend the list with **at least six more** distributions from the
`?Distributions` page, mixing continuous and discrete. Most need
parameters --- check the help page (e.g. `?rgamma`) and pick any sensible
values; skip `rmultinom` (it returns a matrix). Then plot the whole
catalog as histograms in one faceted figure
(`facet_wrap(~ name, scales = "free")`).

The **Cauchy** panel will look broken --- one bar and a huge empty axis.
It is not a bug. Investigate:

- Re-run your `summarize()` of the sample means a few times. Every other
  distribution's mean barely moves between runs. What does the Cauchy's do?
- Look at `sort(abs(rcauchy(n)), decreasing = TRUE)[1:5]` a few times.
  Where is the histogram's x-axis coming from?

**(b)** Pick **two** distributions from your catalog and explore their
parameters, plotting draws across a grid of parameter values. Here is the
pattern, applied to the Normal --- `expand_grid()` builds every
combination, and `map2()` draws a sample for each row:

```{r fig.height = 3}
expand_grid(mean = c(-5, 0, 5), sd = c(0.5, 2)) |>
  mutate(value = map2(mean, sd, ~ rnorm(2000, mean = .x, sd = .y))) |>
  unnest(value) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_grid(sd ~ mean, labeller = label_both)
```

Adapt it to two distributions of your choice (not the Normal). For at
least one of them, find parameter settings that make it match a
distribution you already know.

### Solution

Six additions (any mix works; `?Distributions` has more):

```{r fig.height = 6}
draws <- bind_rows(
  draws,
  list(
    beta    = rbeta(n, 2, 2),
    binom   = rbinom(n, 10, 0.3),
    chisq   = rchisq(n, 3),
    gamma   = rgamma(n, 2),
    pois    = rpois(n, 4),
    weibull = rweibull(n, 2)
  ) |> enframe() |> unnest(value)
)

draws |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ name, scales = "free")
```

The Cauchy investigation:

```{r}
replicate(5, mean(rcauchy(n)))
sort(abs(rcauchy(n)), decreasing = TRUE)[1:5]
```

Every run of `n = 1000` draws contains a few values hundreds of units from
zero, and the histogram's axis stretches to hold them, crushing the bulk
of the data into one bar. The sample mean never settles down, no matter
how often you re-run it --- the Cauchy's tails are so heavy that
$\E{|X|} = \infty$: it **has no mean**, so there is nothing for the sample
mean to converge to. The law of large numbers needs $\E{X}$ to exist, and
here it does not. The Cauchy returns in exercise 6 as the way Monte Carlo
breaks.

For the parameter exploration, the **Beta** is the one to know for this
module --- its two parameters bend it into most shapes a distribution on
$[0, 1]$ can take:

```{r fig.height = 6}
expand_grid(alpha = c(0.5, 1, 2, 5), beta = c(0.5, 1, 2, 5)) |>
  mutate(value = map2(alpha, beta, ~ rbeta(2000, .x, .y))) |>
  unnest(value) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_grid(alpha ~ beta, labeller = label_both)
```

$\text{Beta}(1, 1)$ is exactly $U(0, 1)$ --- the uniform is a special case.
The project's Part B relies on this family, so keep the picture in mind.

And the **Gamma**, whose shape parameter moves it from spiked to
bell-shaped:

```{r fig.height = 2.5}
expand_grid(shape = c(0.5, 1, 2, 10)) |>
  mutate(value = map(shape, ~ rgamma(2000, .x))) |>
  unnest(value) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~ shape, nrow = 1, scales = "free", labeller = label_both)
```

$\text{Gamma}(1)$ is $\text{Exp}(1)$ --- another special case, the same way
the uniform sits inside the Beta family.

---

# 2. An integral on [0, 1]

For an integral over $[0,1]$, $\int_0^1 g(x)\,dx = \E{g(U)}$, where $U$ is uniform on 
$[0, 1]$. We can estimate the value of this integral by taking a sample of IID uniform draws
and computing the sample mean of $g(U_i)$.

## Exercise

Estimate
$$\theta = \int_0^1 \sin(\pi x) \, dx$$
using 10,000 uniform draws. Give a 95\% confidence interval using `t.test`.

The true value is $2/\pi \approx 0.6366$. Does your interval cover it?

### Solution

```{r}
us <- runif(10000)
gs <- sin(pi * us)
mean(gs)
t.test(gs)$conf.int
2 / pi
```

The interval covers $2/\pi$. It should, about 95\% of the time.

---

# 3. Same integral, two RVs

The general recipe works with **any** density $f$ that is positive on the
domain: $\int g(x)\,dx = \E{g(X)/f(X)}$ for $X \sim f$. Different choices
of $f$ give different estimators of the **same** integral. Are they equally
good?

## Exercise

Estimate
$$\theta = \int_0^\infty x^2 e^{-x} \, dx$$
(true value $\Gamma(3) = 2! = 2$) twice, with 10,000 draws each:

**(a)** with $f = $ Exp(1) draws (`rexp` / `dexp`), and

**(b)** with $f = $ Gamma(shape 2, rate 1) draws (`rgamma` / `dgamma`) ---
the Gamma family from exercise 1.

Before coding, simplify $h(x) = g(x)/f(x)$ by hand in each case --- both
simplify pleasantly.

**(c)** Report a 95\% confidence interval for each. Both estimate the same
$\theta$ from the same number of draws --- which interval is narrower, and
by roughly what factor?

**(d)** Plot the **ECDFs** of the 10,000 $h$ values from (a) and the
10,000 from (b) on one set of axes (`stat_ecdf()`). What do the two curves
have in common, and how does their difference explain (c)?

**(e)** Repeat with $f = $ Gamma(shape 3, rate 1). Something
strange happens --- run it, then explain.

### Solution

```{r}
g <- function(x) { x^2 * exp(-x) }
xs1 <- rexp(10000)
h1 <- g(xs1) / dexp(xs1)          # simplifies to xs1^2
xs2 <- rgamma(10000, shape = 2)
h2 <- g(xs2) / dgamma(xs2, 2)     # simplifies to xs2
mean(h1)
mean(h2)
t.test(h1)$conf.int
t.test(h2)$conf.int
```

By hand: $h_1(x) = x^2 e^{-x}/e^{-x} = x^2$, and since the Gamma(2)
density is $x e^{-x}$, $h_2(x) = x^2 e^{-x} / (x e^{-x}) = x$. The
Gamma(2) interval is about **3 times narrower** from the same 10,000
draws: $\Var{X^2} = 20$ under Exp(1) versus $\Var{X} = 2$ under Gamma(2),
and CI width scales with the standard deviation, $\sqrt{20/2} \approx 3.2$.

```{r}
tibble(h = c(h1, h2),
       envelope = rep(c("Exp(1)", "Gamma(2)"), each = 10000)) |>
  ggplot(aes(x = h, color = envelope)) +
  stat_ecdf() +
  coord_cartesian(xlim = c(0, 15))
```

Both ECDFs are centered on the same place --- both estimators are unbiased
for $\theta = 2$ --- but the Exp(1) curve rises slowly with a long tail to
the right, while the Gamma(2) curve climbs steeply. Same target, less
spread per draw, narrower interval. Two estimators of the same quantity
can differ a lot in how much each sample is worth --- the next video calls
this **efficiency**, and spends all its time engineering it.

**(e)** The Gamma(3) density is $x^2 e^{-x} / 2$ --- proportional to the
integrand itself. So
$$h_3(x) = \frac{x^2 e^{-x}}{x^2 e^{-x} / 2} = 2 \quad \text{for every draw:}$$

```{r error = TRUE}
xs3 <- rgamma(10000, shape = 3)
h3 <- g(xs3) / dgamma(xs3, 3)
range(h3)
t.test(h3)$conf.int
```

Every draw returns exactly 2, the variance is zero, and `t.test` refuses
to run on constant data. One draw would have given the exact answer. The
**perfect envelope is proportional to the integrand** --- but writing it
down requires the normalizing constant, which is the integral we wanted.
Getting *close* to proportional without knowing the constant is where the
next video is headed.

---

# 4. Probabilities are expectations of indicators

Recall an indicator function is a function that has the value 1 if a statement
is true and 0 if the statement is false. They are the mathematical equivalent of
`if` statements and conditionals in R. An important result is that for event $A$,

$$P(A) = E(\mathbf{1}(A))$$

In other words, we can estimate probabilities in the same way we estimate expectations.

## Exercise

For $Z \sim N(0, 1)$, estimate
$$\theta = P(1 \le Z \le 2)$$
using 100,000 draws.

1. Write the indicator directly in R (no `ifelse` needed).
2. Give a 99\% confidence interval. Which is the right tool here, `t.test`
   or `binom.test`, and why?
3. Check against the exact value from `pnorm`.

### Solution

```{r}
zs <- rnorm(100000)
hits <- (1 <= zs) & (zs <= 2)
mean(hits)
binom.test(sum(hits), length(hits), conf.level = 0.99)$conf.int
pnorm(2) - pnorm(1)
```

`binom.test` is the right tool: the indicator is a Bernoulli random
variable, so the interval for a proportion applies exactly, not just
through the CLT approximation.

---

# 5. Joint distributions

Nothing in the recipe is one-dimensional. A double integral is an
expectation over a random **pair**: sample both coordinates, apply $h$,
average. When the coordinates are independent, sampling the pair is just
two calls to an `r` function.

## Exercise

**(a)** For $X, Y$ independent $U(0, 1)$, estimate the double integral
$$\theta = \int_0^1 \int_0^1 |x - y| \, dx \, dy = \E{|X - Y|}$$
with 100,000 pairs and a 95\% confidence interval. The true value is $1/3$.

**(b)** For $Z_1, Z_2$ independent $N(0, 1)$, estimate
$$P(Z_1 < Z_2 + 1)$$
with a 99\% confidence interval, using the right interval tool for the
job. Then check it exactly: what is the distribution of $Z_1 - Z_2$?

### Solution

```{r}
xs <- runif(100000); ys <- runif(100000)
diffs <- abs(xs - ys)
mean(diffs)
t.test(diffs)$conf.int
```

```{r}
z1 <- rnorm(100000); z2 <- rnorm(100000)
hits <- z1 < z2 + 1
mean(hits)
binom.test(sum(hits), length(hits), conf.level = 0.99)$conf.int
```

Part (b) is an indicator, so `binom.test`. For the exact check: $Z_1 - Z_2$
is Normal with mean 0 and variance $1 + 1 = 2$, so
$$P(Z_1 - Z_2 < 1) = \Phi(1/\sqrt{2})$$

```{r}
pnorm(1, mean = 0, sd = sqrt(2))
```

Two calls to `rnorm`, and the "double integral" never looked like one. The
same recipe runs in any dimension --- and the video showed its error rate
does not depend on the dimension.

---

# 6. When the recipe breaks

The video stated conditions: $\E{h(X)} < \infty$ **and** $\E{h(X)^2} <
\infty$. This exercise is about what happens when the second one fails.

## Exercise

Estimate
$$\theta = \int_0^\infty \frac{1}{1 + x^2} \, dx$$
using Exp(1) draws, so $h(x) = e^{x} / (1 + x^2)$. The true value is
$\pi/2 \approx 1.5708$.

1. Run the estimator with $n = 100{,}000$ **five separate times**
   (`replicate` helps). Do the answers agree with each other? With the
   truth?
2. The estimator is unbiased --- we proved that in the video. So what is
   going wrong? Look at $\E{h(X)^2}$.
3. Would increasing $n$ fix this?

### Solution

```{r}
h <- function(x) { exp(x) / (1 + x^2) }
replicate(5, mean(h(rexp(100000))))
pi / 2
```

The five runs disagree wildly, and typically all sit **below** the truth,
with occasional runs far above it.

The variance condition fails:
$$\E{h(X)^2} = \int_0^\infty \frac{e^{2x}}{(1+x^2)^2} \, e^{-x} \, dx
= \int_0^\infty \frac{e^{x}}{(1+x^2)^2} \, dx = \infty$$
since $e^x$ beats any polynomial. Rare draws far in the tail carry
astronomically large $h$ values: most runs miss them (estimate too low) and
an occasional run catches one (estimate far too high). With infinite
variance the CLT does not apply, so the confidence interval machinery is
**invalid** --- the intervals are narrow and wrong.

Increasing $n$ does not fix it. The problem is the mismatch between where
$1/(1+x^2)$ has its mass (heavy tail) and where Exp(1) samples (light
tail). The fix is a better sampling distribution --- that is the subject of
the next video (importance sampling).

---
