---
title: "Monte Carlo Integration"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
institute: "Computational Methods in Statistics and Data Science (DATASCI 406)"
date: "Module 1: Integrate"
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: false
    includes:
      in_header: ../preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%")
options(width = 60, digits = 4)
set.seed(49482)
library(tidyverse)
```

# Integrals as Expectations

## Expectation

Suppose we are going to compute $g(X)$ for a random variable $X$.

We often want to "average over" $X$ to get a sense of a typical value for $g(X)$. We define the **expectation** of $g(X)$ as:
\[
\begin{aligned}
E(g(X)) &= \sum_{i=-\infty}^\infty P(X = x) g(x) & \text{(discrete)}\\
E(g(X)) &= \int_{-\infty}^\infty g(x) f(x) \, dx & \text{(continuous)}
\end{aligned}
\]
We call the **expectation of the identity function**, $E(X)$ the **mean of $X$**.

We call $E\left[(X - E(X))^2\right]$ the **variance**.

## Expectation and Inference

Recall, a **parameter** is an **unknown quantity** in a statistical model.

We previously discussed two forms of **statistical inference** for parameters:

- **Estimation**: making informed guesses about population values.
- **Testing**: checking if the data conform to a specific value of the parameter.

Expectations are useful for both:

- Important **operating characteristics** of estimators (e.g. bias)
- Computing **Type I and Type II** error rates of tests
- Many parameters can be **expressed as expectations**.

## Monte Carlo Integration

Recall your introduction performing integrals using **Riemann sums**:
$$\int_a^b h(x) \, dx \approx \sum_{i = 0}^n  h(a + di/2) \times d, \quad d = \frac{b - a}{n}$$
(or using the trapezoid rule or any other similar technique).

Straightforward, but

- How do you pick $d$? Alternatively, if spacing is unequal, how do you pick the regions?
- When integrating in $k$-dimensions we need to take $n^k$ samples (gets big fast!)

Solution: let $h(x)$ help us pick the most important regions and integrate using randomly selected points.

## Example: Universal function integrator

Suppose we want to evaluate a complex function over the region 0 to 1 (for a math class).

For example,
$$\int_0^1 \left(\log\left(\frac{1}{x}\right)\right)^3 \, dx$$

This integral doesn't have a **closed form solution**, so our usual techniques do not work. (BTW: This is the $\Gamma(4)$ function.)

## Example continued

Recall: if $U \sim U(0,1)$ then the density function is $f(u) = 1$.

$$\int_0^1 \left(\log\left(\frac{1}{x}\right)\right)^3 \, dx = \int_0^1 \left(\log\left(\frac{1}{x}\right)\right)^3 f(x) \, dx  = E\left(g(U)\right)$$
where
$$g(X) =\left(\log\left(\frac{1}{x}\right)\right)^3$$
Conveniently, computers are great at generating lots of $U(0,1)$ random variables!

We'll **approximate $E(g(U))$** with **the sample mean of $g(U_i)$**.

## Monte Carlo Gamma Function

We **estimate the integral** using draws from $U(0,1)$ and the sample mean of the function values:

```{r }
g <- function(u) { log(1/u)^3 }
mean(g(runif(1000000)))
```

Since the exponent was integer, $\Gamma(a) = (a - 1)!$, in this case **3! = 6**.

## Monte Carlo integration: the general case.

We wish to estimate $\int_a^b g(x) \, dx$. Suppose $X$ is a random variable with:

- Support $[a, b]$ (where either $a \rightarrow -\infty$ or $b \rightarrow \infty$)
- Density $f(x)$ that is non-zero over $[a, b]$

then
$$\int_a^b g(x) \, dx = \int_a^b g(x) \frac{f(x)}{f(x)} \, dx  = \int_a^b \frac{g(x)}{f(x)} f(x) \, dx  = \E{h(X)}$$
    where $h(x) = g(x) / f(x)$.

Sample from $X$ i.i.d. and estimate using $(1/n) \sum_{i=1}^n g(X_i)/f(X_i)$.

## Example: Integral of $1/2^x$

Suppose we need to compute
$$\int_0^\infty \frac{1}{2^x} \, dx$$

What kind of distribution has the support $[0, \infty)$?

The **exponential distribution** has

- Support $[0, \infty)$
- Density function $f(x) = \exp\{-x\}$ (keeping the usual parameter $\lambda = 1$)
- Random number generator `rexp`

## Example continued

We've identified a variable on $[0, \infty)$, what expectation should we estimate?

Let $g(x) = 1/2^x$.
$$\int_0^\infty g(x) \, dx = \int_0^\infty \frac{g(x)}{\exp\{-x\}} \exp\{-x\} \, dx = \E{h(X)}$$

```{r echo=FALSE}
set.seed(393921292)
```

## Implementing

```{r }
g <- function(x) {
  1/(2^x)
}
h <- function(x) { g(x) / dexp(x) } ## R's exp. density function
```

```{r }
k <- 100000
hX <- h(rexp(k))
mean(hX)
```

```{r }
1/log(2)
```

## Distributions in R

R has a naming convention for functions related to **distributions** based on **prefixes**:

- `r`: generated random values from the distribution (deviates)
- `d`: evaluates the probability density or mass function at its argument
- `p`: implements the CDF for $P(X \le x)$
- `q`: implements the quantile function (inverse of CDF) and finds $x$ such that $P(X \le x) = p$.

See `?Distributions` for a list of built in distributions with these functions.

## Monte Carlo: Special case

Sometimes the quantity we want **is already an expectation**:
$$E(g(X)) = \int g(x) f(x) \, dx$$
for a random variable $X$ we know how to sample. Then there is **no trick at all**: sample from $X$ and average $g(X_i)$.

Example: $X \sim \text{Exp}(1)$, estimate $E\left(\log(1 + X)\right)$:

```{r }
mean(log(1 + rexp(100000)))
```

The density $f$ never appears in the code — it is baked into `rexp`.

# Why It Works

## Estimating a mean (in general)

Why does this work? Suppose that $X_i \iid F$ and $E(g(X)) < \infty$. The **sample mean of $g(X_i)$** is **unbiased** for the population quantity **$E(g(X))$**.

\[
\begin{aligned}
E\left(\frac{1}{n} \sum_{i=1}^n g(X_i)\right) &= \frac{1}{n} \sum_{i=1}^n E(g(X_i)) \\
&= \frac{1}{n} n E\left(g(X_1)\right) \\
&= E(g(X))
\end{aligned}
\]

Provided we can sample from $F$, **the sample mean is a good estimator of $E(g(X))$**.

## Larger sample $\rightarrow$ better estimates

```{r }
g <- function(u) { log(1/u)^3 }
ms <- map_dbl(2:20, ~ mean(g(runif(2^.x))))
```

```{r echo=FALSE, fig.height=2.5, fig.width=6}
par(mar = c(4,4,0,0))
plot(ms, xlab = "Sample Size (log_2)", ylab = "Estimate", type = 'b')
abline(h = 6, col ="red", lty = 2)
```

## Convergence of Estimates

When $\E{g(X)} < \infty, \E{g(X)^2} < \infty$ and $X_i$ are IID, by the **weak law of large numbers**,
$$\hat \theta_n = \frac{1}{n} \sum_{i=1}^n g(X_i) \overset{P}{\rightarrow} \E{g(X)} = \theta$$
as $n \rightarrow \infty$.

Reminder: The notation $\overset{P}{\rightarrow}$ indicates "convergence in probability":
$$\hat \theta_n \overset{P}{\rightarrow} \theta \equiv \lim_{n\rightarrow\infty} \Pr(|\hat \theta_n - \theta| < \epsilon) = 1, \quad \text{for any } \epsilon > 0$$
In words: we can draw a large enough sample to concentrate all the probability of $\hat \theta_n$ within $\pm \epsilon$ of $\theta$.

## Uncertainty in estimating $\E{g(X)}$

Under these same conditions ($X_i \iid F, \E{g(X)} < \infty, \E{g(X)^2} < \infty$), the **central limit theorem** states:
$$\frac{\hat \theta_n - \theta}{\hat s_n/\sqrt{n}} \overset{D}{\rightarrow} N(0, 1) \quad \text{(convergence in distribution)}$$
where:
$$\hat \theta_n = \frac{1}{n} \sum_{i=1}^n g(X_i), \quad \hat s_n = \sqrt{\frac{1}{n - 1} \sum_{i=1}^n (g(X_i) - \hat \theta_n)^2}$$

## Confidence Intervals for $\theta$

Our Monte Carlo answer is an **estimate**, not an exact solution, so we report a **confidence interval** (a plausible range) for the value of the integral.

By the CLT,
$$\hat \theta_n \approx N(\theta, \tau^2/n), \quad \tau^2 = \text{Var}(g(X))$$
an approximate $100 \times (1 - \alpha) \%$ **confidence interval for $\theta$** is given by:
$$\hat \theta \pm t_{\alpha/2}(n - 1) \times \hat s / \sqrt{n} $$
where $t_{\alpha/2}(n - 1)$ is the $\alpha/2$ quantile of Student's $t$-distribution with $n - 1$ degrees of freedom.

## Confidence Intervals in R

To estimate
$$\int_0^\infty \frac{1}{2^x} \, dx$$
we used an exponential random variable so generate the estimate:

```{r }
mean(hX)
```

To add a 99.9\% confidence intervals:

```{r }
## t.test will do other things, we only need the conf.int
t.test(hX, conf.level = 0.999)$conf.int
```

## Confidence Intervals with Larger Samples

```{r echo=FALSE, fig.height=3, fig.width=6}
js <- seq(100, length(hX), length.out = 100) %>% round
cis <- sapply(js, function(j) { t.test(hX[1:j], conf.level = 0.999)$conf.int})
tmp <- data.frame(t(cis))
colnames(tmp) <- c("lower", "upper")
tmp$size <- js
ggplot(tmp, aes(x = size)) + geom_line(aes(y = lower)) + geom_line(aes(y = upper))
```

# Probabilities and CDFs

## Probability as an Integral

Many statistical tasks can be viewed as taking an expectation for a suitable function $g$.

In particular, if $g$ is an **indicator function**, then the expectation of $g$ is a **probability**.

**Indicator functions**:
\[
\ind(s) = \left\{ \begin{array}{lc} 1 &: s \text{ is true} \\ 0 &: s \text{ is false} \end{array} \right.
\]

Note: In R we get indicators "for free" by writing things like: `x <= 3`.

## Expectation of Indicators

Suppose we have a **continuous random variable $X$** and **$g$ be an indicator function**. For example, $g(X) = \ind(X > 3)$.

Since $g(X)$ can be either 1 or 0 (it is a **random variable**), let **$A$ be the event** (values of $X$) that $g(X)$ is 1.

**Notice**: $A$ and $A^c$ partition the sample space.

**Claim**: $E(g(X)) = P(A)$

Example: $E(\ind(X  > 3)) = P(X > 3)$

## Proof

Remember: $g(X) = 1$ if $A$ is true and $g(X) = 0$ if $A^c$ is true ("not $A$").
\[
\begin{aligned}
    E(g(X)) &= \int_{-\infty}^\infty g(x) f(x) \, dx  \\
    &= \int_A  1 \times f(x) \, dx  + \int_{A^c} 0 \times f(x) \, dx  \\
    &= \int_A f(x) \, dx  \\
    &= P(A)
\end{aligned}
\]

## Estimating a Cumulative Distribution Function

The CDF of a (continuous) random variable is defined as:
$$F(t) = P(X \le t) = E(\ind(X \le t))$$

The **empirical CDF** is then the sample mean of $\ind(X_i \le t)$:
$$\hat F_n(t) = \frac{1}{n} \sum_{i=1}^n \ind(X_i \le t)$$

This is the **first case** again: the density is already inside the integral, so we sample $X$ directly and average — no multiply/divide needed.

## Example: Normal CDF

The CDF of the Normal distribution **does not have a closed form**.

For $X \sim N(0, 1)$, let's estimate $P(X \le 1.96)$.

```{r }
xs <- rnorm(1e5)
mean(xs <= 1.96)
```

##

```{r echo=FALSE, fig.height=3.5, fig.width=6}
par(mar = c(4, 4, 2, 0))
grid <- seq(-2, 2, length.out = 200)
plot(grid, sapply(grid, function(g) { mean(xs <= g)}), type = 'l', xlab = "x", ylab = "F(x)")
curve(pnorm(x), add = TRUE, col = "red", lty = 2, lwd = 3)
```

## Better Confidence Intervals for Estimated Probabilities

The **$t$-distribution intervals** we used earlier would work when estimating a probability, but **we can do slightly better**.

Consider the random variable $W = \ind(X \le t)$. **What is this?**

- $W$ can take values of 0 or 1
- $P(W = 1) = P(X \le t) = \theta$, $P(W = 0) = 1 - \theta$.
- $W$ is **Bernoulli**!

We have to estimate $\theta$ (our goal anyway), but this means we can use **confidence intervals for proportions** (i.e., `binom.test` instead of `t.test`).

```{r }
binom.test(sum(xs <= 1.96), n = length(xs))$conf.int
```

# Higher Dimensional Problems

## Integrals in more than one dimension

Nothing in our recipe was one-dimensional. To integrate over a region of $\mathbb{R}^k$: sample a random **vector**, apply $g$, average.

The classic double integral — the area of the unit circle:
$$\pi = \int_{-1}^{1} \int_{-1}^{1} \ind(x^2 + y^2 \le 1) \, dx \, dy$$

Take $X, Y \iid U(-1, 1)$, so the joint density is $f(x, y) = 1/4$ on the square. Multiply and divide:
$$\pi = 4 \int_{-1}^{1} \int_{-1}^{1} \ind(x^2 + y^2 \le 1) \, \tfrac{1}{4} \, dx \, dy = 4 \, \E{\ind(X^2 + Y^2 \le 1)}$$

## Throwing darts at a square

```{r }
n <- 100000
x <- runif(n, -1, 1); y <- runif(n, -1, 1)
inside <- x^2 + y^2 <= 1
4 * mean(inside)
4 * binom.test(sum(inside), n)$conf.int
```

An indicator again, so `binom.test` gives the confidence interval — everything from the last section carries straight over.

##

```{r echo=FALSE, fig.height=3, fig.width=6}
par(mar = c(2, 2, 0.5, 0.5), pty = "s")
show <- 1:3000
plot(x[show], y[show], col = ifelse(inside[show], "red", "grey60"),
     pch = 16, cex = 0.3, xlab = "", ylab = "", asp = 1)
theta <- seq(0, 2 * pi, length.out = 400)
lines(cos(theta), sin(theta), lwd = 2)
```

## Aside: big-O notation

The CLT says the Monte Carlo error shrinks like $1/\sqrt{n}$. Is that fast?

To compare methods, we describe **how fast the error shrinks** as $n$ grows:
$$\text{error} = O\left(g(n)\right)$$
means: for some constant $C$, $|\text{error}| \le C \, g(n)$ for all large $n$.

The constant is deliberately ignored — big-O records the **rate**, which dominates once $n$ is large.

A better rate always wins *eventually* — the question is how large $n$ has to be first.

## MC Error Rate $O(n^{-1/2})$

Recall the **standard deviation of a sample mean**:
$$\Var{\hat \theta_n} = \Var{\frac{1}{n} \sum_{i=1}^n g(X_i)} = \frac{1}{n^2} \sum_{i=1}^n \Var{g(X_i)} = \frac{\tau^2}{n}, \quad \tau^2 = \Var{g(X)}$$

so the typical error is
$$\sqrt{\tau^2 / n} = \tau \, n^{-1/2}$$

The population variance $\tau^2$ sets the **constant**; $n$ sets the **rate**. So the error rate of Monte Carlo is $O(n^{-1/2})$. Notice that this **does not depend on $k$** -- though the constant $\tau$ typically grows with $k$.

## Grid search error rate $O(n^{-2/k})$

Take one cell of width $d$ with midpoint $m$, and Taylor expand the integrand:
$$h(x) = h(m) + h'(m)(x - m) + \tfrac{1}{2} h''(\xi) (x - m)^2$$

- Integrating over the cell, the **linear term vanishes**: $(x - m)$ is symmetric about the midpoint. 
- The leftover per cell is $\tfrac{1}{2} h'' \int_{\text{cell}} (x - m)^2 \, dx \approx h'' \, d^3 / 24$.
- Summing over the $(b - a)/d$ cells: total error $\approx (b - a) \, \overline{h''} \, d^2 / 24 = O(d^2)$.
- With $n$ points, $d = (b - a)/n$, so the error is $O(n^{-2})$.
- In $k$ dimensions, $n$ points make the cells $d \propto n^{-1/k}$ wide: error $= O(d^2) = O(n^{-2/k})$.

## Comparison using larger samples, more grid points.

```{r echo=FALSE, fig.height=2.6, fig.width=6}
par(mar = c(4.2, 4, 0.5, 7.5), xpd = NA)
n <- 10^seq(2, 6, length.out = 200)
plot(n, n^(-1/2), log = "xy", type = "l", lwd = 3, col = "red",
     ylim = c(1e-12, 1), xlab = "n (total function evaluations)",
     ylab = "error (up to a constant)")
for (k in c(1, 4, 20)) lines(n, n^(-2/k), lty = 2)
text(2e3, 1.8 * (2e3)^(-1/2), "Monte Carlo (every k)", col = "red", adj = c(0, 0))
text(1.4e6, 1e6^(-2/20), "grid, k = 20", adj = 0)
text(1.4e6, 1e6^(-2/4), "grid, k = 4 (tie)", adj = 0)
text(1.4e6, 1e6^(-2/1), "grid, k = 1", adj = 0)
```

Most real world problems are high dimensional (like linear regression). For this class we will tend to work with simpler examples.

## Putting it to the test: a high-dimensional CDF

We just estimated the Normal CDF in one dimension. Now the **joint CDF** of $Z_1, \ldots, Z_k \iid N(0,1)$, evaluated at $(1, \ldots, 1)$:
$$\theta_k = F(1, \ldots, 1) = P\left(Z_1 \le 1, \ldots, Z_k \le 1\right) = \int_{-\infty}^{1} \!\!\cdots\! \int_{-\infty}^{1} \prod_{j=1}^k \phi(z_j) \, dz = \Phi(1)^k$$
The exact answer is known, so we can score both methods. **Same budget for both**: $10^5$ function evaluations.

- Grid: needs a *bounded* box first — chop at $-8$, where $\Phi(-8) \approx 6 \times 10^{-16}$. Then $m = \lfloor (10^5)^{1/k} \rfloor$ midpoints per axis on $[-8, 1]^k$.
- Monte Carlo: $10^5$ draws of $Z \in \mathbb{R}^k$ — no box needed.

(Because the integrand factors, the $k$-dimensional midpoint value is just the 1-D midpoint value to the $k$-th power — we can *score* the full grid without enumerating it.)

## The code

```{r }
exact1 <- pnorm(1)  # one axis: P(Z <= 1)
n <- 1e5
grid_err <- map_dbl(1:10, function(k) {
  m <- floor(n^(1 / k) + 1e-9)               # points per axis
  mids <- seq(-8 + 4.5/m, 1 - 4.5/m, length.out = m)
  I1 <- sum(dnorm(mids)) * (9 / m)           # 1-D midpoint rule
  abs(I1^k - exact1^k)                       # grid value is I1^k
})
mc_err <- map_dbl(1:10, function(k) {
  zs <- matrix(rnorm(n * k), ncol = k)
  abs(mean(rowSums(zs <= 1) == k) - exact1^k)
})
```

##

```{r echo=FALSE, fig.height=3, fig.width=6}
par(mar = c(4.2, 4, 0.5, 1))
ks <- 1:10
mc_se <- sqrt(exact1^ks * (1 - exact1^ks) / n)
plot(ks, grid_err, log = "y", type = "b", lty = 2, pch = 1,
     ylim = range(c(grid_err, mc_err, mc_se)),
     xlab = "dimension k", ylab = "|error|")
points(ks, mc_err, col = "red", pch = 16)
lines(ks, mc_se, col = "red", lwd = 2)
legend("bottomright", c("grid (midpoint rule)", "Monte Carlo (one run)", "Monte Carlo std. error"),
       pch = c(1, 16, NA), lty = c(2, NA, 1), col = c("black", "red", "red"), bty = "n", cex = 0.8)
```

<!--
## The arithmetic at $k = 20$

Even a coarse grid, 10 points per axis, in 20 dimensions: $10^{20}$ evaluations.

At a billion evaluations per second, that is about **3,000 years**.

Monte Carlo gets 3 digits from $10^6$ draws in **seconds** — in any dimension.

Bonus: the grid must be laid down **before** you know where $f$ puts its mass; sampling from $f$ concentrates the evaluations there automatically.

## Checking it in $k = 10$

For $Z_1, \ldots, Z_{10} \iid N(0,1)$, estimate $P(\max_j |Z_j| \le 1)$ — which we can verify exactly: $(2\Phi(1) - 1)^{10}$.

```{r }
k <- 10; n <- 1e6
zs <- matrix(rnorm(n * k), ncol = k)
mean(rowSums(abs(zs) <= 1) == k)
(2 * pnorm(1) - 1)^10
```

A ten-dimensional integral, one million draws, three matching digits.
-->

## Monte Carlo Integration: Summary

- Monte Carlo turns an integral into a **sampling problem**.
- To estimate $\int_a^b g(x) \, dx$: pick a density $f$ positive on $[a, b]$, sample $X_i \iid f$, average $g(X_i)/f(X_i)$.
- If the target is already $\E{g(X)}$: sample and average, no ratio needed.
- R functions for RVs: r` samples, `d` is the density (`p`, `q` for CDF and quantiles).
- The sample mean is **unbiased** and **consistent** (LLN); the CLT gives **confidence intervals** using `t.test`.
- Probabilities are expectations of **indicators**, but use `binom.test` for CI.
- Multiple dimensions: sample a random **vector**, apply $g$, average.
- Error is $O(n^{-1/2})$ **in every dimension**; grids are $O(n^{-2/k})$, poor for large $k$.

