---
title: "Variance Reduction and Importance Sampling: Session Exercises"
subtitle: "Module 1, Day 2 --- DATASCI 406, Computational Methods in Statistics and Data Science"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 3.5,
                      collapse = TRUE, comment = "#>")
library(tidyverse)
set.seed(40602)
```

\newcommand{\E}[1]{\text{E}\left(#1\right)}
\newcommand{\Var}[1]{\text{Var}\left(#1\right)}
\newcommand{\ind}{\mathbf{1}}

---

# 1. Antithetic variables without uniforms

The video built antithetic pairs from uniforms, because $U$ and $1 - U$ have
the same distribution. But the idea is not about uniforms: it needs only
some way to produce a second draw with **the same distribution** that is **negatively correlated**.

For any *symmetric* distribution, negation is an easy way to achieve this result. If
$Z \sim N(0,1)$ then $-Z \sim N(0,1)$ too, so $(Z, -Z)$ is an antithetic
pair, and for monotone $h$ the pair $(h(Z), h(-Z))$ is negatively
correlated.

We will estimate
$$\theta = \E{e^Z} = \int_{-\infty}^{\infty} e^z \phi(z) \, dz = e^{1/2} \approx 1.6487$$

## Exercise

**(a)** Estimate $\theta$ two ways with the same 10,000 evaluations of $h$:
10,000 IID `rnorm` draws ($Z_i$), and 5,000 antithetic pairs
$[e^{V_i} + e^{-V_i}]/2$. Report both estimates, the estimated variance of
each *estimator*, and the percent variance reduction.

Then compute `cor(exp(v), exp(-v))` on your 5,000 draws and put it next to
the percent reduction. What do you notice?

**(b)** A classmate objects that the pairing is only bookkeeping: the
estimate is the average of all 10,000 values either way, so why not pool
them and be done?

```{r eval = FALSE}
# remove eval = FALSE to run this chunk after creating v and antithetic pairs
pooled <- c(exp(v), exp(-v))
c(pooled = mean(pooled),           paired = mean(anti))
c(pooled = var(pooled) / 10000,    paired = var(anti) / 5000)
```

One of those two comparisons agrees exactly and the other does not. Which
variance is telling the truth? Settle it **empirically**: repeat the entire
procedure 2,000 times and compare the actual variance of the 2,000 estimates
against both candidates. The `replicate` function will be helpful.

**(c)** Now try $h(z) = z^2$, so $\theta = \E{Z^2} = 1$. Something goes
badly wrong. Why does that antithetic technique not work here?

---

# 2. Rare events

Direct sampling is a bad way to estimate a small probability, because
almost every draw contributes a zero. Importance sampling moves the
sampling distribution to where the action is and reweights.

## Exercise

Estimate $\theta = P(Z \ge 3.5)$ for $Z \sim N(0,1)$; the exact value is
`pnorm(3.5, lower.tail = FALSE)`.

**(a)** Direct: 10,000 draws from `rnorm`, average the indicators, and give
a 95% interval with the right tool. How many draws landed in the region?
What are the only values this estimator can possibly return?

**(b)** Importance sampling with a shifted exponential: draw
$Y = 3.5 + \text{Exp}(\lambda)$, so
$$g(y) = \lambda e^{-\lambda (y - 3.5)}, \quad y > 3.5,$$
and every draw is in the tail. Then
$$\theta = \E{\ind(Y \ge 3.5) \frac{\phi(Y)}{g(Y)}} = \E{\frac{\phi(Y)}{g(Y)}}.$$
Run this with $\lambda = 1$ and 10,000 draws. Compare the estimated
variance of the estimator with the exact variance of (a)'s.

**(c)** Try $\lambda \in \{1, 2, 3, 4, 5, 8\}$. Which is best, and why is
it not $\lambda = 1$? (Hint: what does the ideal envelope from the video
look like here, and what is the shape of the normal density just past
3.5?) Also print `range()` of the weights for your best $\lambda$ --- what
does the video's safety condition say about them?

---

# 3. Which direction of tail mismatch is safe?

Exercise 2 ran into trouble with an envelope that was too *light* in the
tail. This settles the general question with a matched pair of runs.

## Exercise

The Cauchy has heavy (polynomial) tails; the Normal has light
(exponential-in-$x^2$) tails. Estimate a tail probability in each
direction, five runs of 100,000 draws each:

**(a)** target Cauchy, envelope Normal: $P(C \ge 2)$ using `rnorm` draws.

**(b)** target Normal, envelope Cauchy: $P(Z \ge 2)$ using `rcauchy` draws.

One of these is fine and the other is not. Which, and state the general
rule in one sentence. Then explain it with the finite-variance condition
$\E{h(X)^2 f(X) / g(X)} < \infty$ from the video.

---
