---
title: "Benford's Law: Session Exercises"
subtitle: "Module 1, Day 5 --- DATASCI 406, Computational Methods in Statistics and Data Science"
author: "Mark M. Fredrickson (mfredric@umich.edu)"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 3.5,
                      collapse = TRUE, comment = "#>")
library(tidyverse)
set.seed(40605)
```

**Benford's Law** says the leading digit of a number --- the 2 in 242, the 5
in 5670 --- is not uniform on $\{1, \ldots, 9\}$ but follows

$$P(D = d) = \log_{10}\!\left(\frac{d+1}{d}\right), \qquad d = 1, \ldots, 9,$$

so ones turn up about 30% of the time and nines about 4.6%. It is used in
tax auditing and election forensics on the theory that fabricated numbers do
not obey it.

---

# Leading Digits

Let $x$ be any real number with $x \ge 1$ and write:

$$x = 10^{m + f}$$

Where $m$ is the largest integer such that $x \ge 10^m$. Then by construction
$m \ge 0$ and $0 \le f < 1$. Splitting $x$
this way separates its **scale** from its **digits**:

$$x = 10^m \cdot 10^f.$$

Because $0 \le f < 1$ we have $1 \le 10^f <
10$, and multiplying by $10^m$ only moves the decimal point. So $10^f$ is the
digit string of $x$ with the point placed after the first digit --- the
**mantissa**. For $242 = 10^2 \cdot 2.42$ it is $2.42$; for
$5670 = 10^3 \cdot 5.67$ it is $5.67$. The leading digit is therefore

$$D(x) = \left\lfloor 10^{f} \right\rfloor, \qquad f = \log_{10}(x) \bmod 1.$$

Due to the finite precision of computers, we have to compute $D(x)$ slightly differently than the ideal form.
Rearranging $x = 10^m \cdot 10^f$ gives $10^f = x / 10^m$, so the same digit
is equally

$$D(x) = \left\lfloor \frac{x}{10^{m}} \right\rfloor,
\qquad m = \lfloor \log_{10} x \rfloor.$$

## Exercise 1 --- Computing the first digit

Write `first_digit(x)` using the second form. The floor operator is `floor()`.

```{r}
# solution
```

Check your solution:

```{r eval = FALSE}
first_digit(c(242, 5670, 91000))
first_digit(c(1:9, 80, 900))
```

# Distributions over $D$

Notice that many values for $f$ will result in the same first digit because we take
the floor of $10^f$. A digit therefore corresponds not to one value of $f$ but to a
whole **set** of them --- every $f$ with $\lfloor 10^f \rfloor = d$, which is
every $f$ satisfying

$$d \; \le \; 10^{f} \; < \; d + 1.$$

The set of $f$ that produce digit $d$ is an **interval**:

$$I_d = \left[\, \log_{10} d, \; \log_{10}(d+1) \,\right) \qquad d = 1, \ldots, 9$$

A distribution over leading digits is therefore a distribution over the intervals $I_d$. To generate random numbers for this distribution, imagine picking an interval $I_d$, then picking an $F$ within that interval. Select $M$ following some other method and $X = 10^M 10^F$.

## Exercise 2 --- Uniform leading digits

Generate 10,000 draws this way: choose one of the nine intervals $I_d$
uniformly, choose $F$ uniformly within it, choose $M \sim \text{Poisson}(2)$,
and set $X = 10^{M} 10^{F}$. Plot a histogram of the draws.

# Benford's Law 
The interval $I_d$ for digit $d$ has width

$$w_d = \log_{10}(d+1) - \log_{10} d = \log_{10}\!\left(\frac{d+1}{d}\right),$$

and these widths are nowhere near equal. Digit 1 owns 30% of the interval (0,1), digit
9 owns 4.6%:

```{r}
# widths
round(log10((2:10) / (1:9)), 3)
```

Benford's Law states that the probability of selecting $F$ to be in $I_d$ is equal to $w_d$. A sufficient condition to be true is when $F$ is uniform on (0,1)

## Exercise 3 --- Benford leading digits

Repeat the previous exercise using $F \sim U(0,1)$

# Real World Benford

So why would we expect the distribution over the intervals $I_d$ to be equal to their widths, or that $F$ was uniform on $(0,1)$?

Recall the Central Limit Theorem states that for IID variables $Y_i$ with mean $\mu$ and variance $\sigma^2$, $\sum_{i=1}^k Y_i \approx N(k \mu, k \sigma^2)$. In this particular case, suppose that $\mu > 0$. Write $V = 10^Y$, so that $\sum_{i=1}^k \log_{10} V_i \approx N(k \mu, k \sigma^2)$.

Split that sum into its integer and fractional parts,
$\sum_{i=1}^k \log_{10} V_i = M + F$, as we did in the last section. When $k$ is large, the mean of $M + F$ will be $k \mu$, far from zero, and the variance $k \sigma^2$ will imply a very flat distribution over many units. With enough spread $F$ becomes
uniform on $(0,1)$, and $P(F \in I_d) \approx w_d$.

Now reverse this logic: $10^{M+F} = 10^{\sum_i \log_{10} V_i}.$ Recall, a sum of logs is the log of a product: 
$$10^{\sum_i \log_{10} V_i} = 10^{\log_{10} \prod_i V_i} = \prod_i V_i$$
Writing $X = \prod_{i=1}^k V_i$ we can see that any
quantity built by multiplying many IID factors together has Benford leading
digits.

## Exercise 4 --- Products of $k$ factors

Let $Y_i \sim \text{Exp}(10)$, so every $V_i = 10^{Y_i}$ is at least 1, and
let $X = \prod_{i=1}^k V_i$. For $k \in \{1, 5, 25, 100\}$ simulate 100,000
values of $X$, plot the distribution of $F = \log_{10} X \bmod 1$,and tabulate the leading digits
against $w_d$.

How large does $k$ have to be?

# A test on real data

Tam Cho and Gaines report leading-digit frequencies for **8,396
contributions between political committees**, from 2004 FEC filings (their
Table 1):

```{r}
pol_digits <- c(23.3, 21.1, 8.5, 11.7, 9.5, 4.2, 3.7, 4.0, 14.1) / 100
```

```{r echo = FALSE, fig.height = 2.5}
tibble(d = factor(rep(1:9, 2)), p = c(pol_digits, log10(1 + 1 / 1:9)),
       source = rep(c("FEC contributions", "Benford"), each = 9)) |>
  ggplot(aes(x = d, y = p, fill = source)) + geom_col(position = "dodge") +
  labs(x = "leading digit", y = NULL, fill = NULL)
```

They disagree, most visibly at 9: 14.1% against an expected 4.6%. Is that
surprising, or is it what random data looks like?

Instead of $\chi^2$, use the Euclidean distance between the observed and
expected digit distributions,

$$T = \sqrt{\sum_d \left(\hat p_d - p_d\right)^2}.$$

We write it $T$ because it is going to be a test statistic --- and because $m$
is already taken, by the integer exponent in the opening derivation.

This is the paper's own statistic. Tam Cho and Gaines report it divided by the
largest distance attainable --- all mass on digit 9, which gives 1.036 --- and
call that $d^*$. So the 0.131 in their Table 1 for 2004 is the $T$ you are
about to compute, rescaled. They chose a distance precisely *because* it is not
tied to a hypothesis test and does not move with sample size; Exercises 6 and 7
then build a $p$-value and a power curve on it.

These do the bookkeeping. The only piece left to write is `distance()`.

```{r}
dbenford <- function(d) log10((d + 1) / d)
digit_props <- function(d) as.numeric(table(factor(d, levels = 1:9))) / length(d)
rbenford <- function(n) sample(1:9, n, prob = dbenford(1:9), replace = TRUE)
tstat <- function(d) distance(digit_props(d))
```

## Exercise 5 --- The observed statistic

Write `distance(v)`, taking nine proportions and returning $T$. Compute the
observed value for `pol_digits`.

## Exercise 6 --- The null distribution and the $p$-value

Under $H_0$ the 8,396 digits are iid draws from the Benford distribution.
Simulate the null with `B = 4000` replicates, plot it with the observed value
marked, and compute the $p$-value with the $(1 + \#)/(B+1)$ formula.

A classmate reports "$p = 0$, so the data cannot be Benford." What is wrong
with that, and what should be reported instead?

## Exercise 7 --- Power against a flatter alternative

Suppose the truth is not Benford but something more uniform --- halfway
between the two:

```{r}
p_alt <- 0.5 * dbenford(1:9) + 0.5 / 9
round(p_alt, 3)
```

For each $n \in \{50, 200, 8396\}$: simulate 1,000 null statistics at that
$n$, take the 95% quantile as the critical value, simulate 1,000 more with
digits drawn from `p_alt`, and report the fraction above the cutoff. That
fraction is the power against this alternative.

What does the $n = 50$ row say about a Benford test run on a small dataset?

---
