---
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

**DATASCI 406 FA 26, Quiz 1.5**

**Name and Uniqname:**

# Question 1 (4 points)

Using today's exercises, identify an example of each of these techniques we
have covered in this module.

**(a)** Estimating a feature of a distribution (such as an expectation or
probability) for which you do not have a closed form answer.

<!-- answer-space: 0.85in -->

**(b)** Finding the operating characteristic of an estimator or testing
procedure.

<!-- answer-space: 0.85in -->

**(c)** Approximating the sampling distribution of a statistic that has no
tractable form.

<!-- answer-space: 0.85in -->

**(d)** Investigating a claim you have no formal proof of.

<!-- answer-space: 0.85in -->

### Solution

**(a)** The $p$-value.

**(b)** The power against `p_alt`.

**(c)** The 4,000 simulated distance statistics --- the null distribution.

**(d)** That $F$ becomes uniform as $k$ grows, or that $F \sim U(0,1)$ gives
digits at $w_d$.

**Rubric.** 1 pt each, generously. Any exercise fitting the description
counts, and one exercise may serve more than one part --- the null simulation
is a fair answer to both (a) and (c).

# Question 2 (1 point)

The paper's conclusion mentions tests built on the **second** digit rather
than the first. To run today's test that way you would need the distribution
of the second digit under the null.

Assume Benford's Law still holds --- that is, $F \sim U(0,1)$. Explain how you
would use a Monte Carlo technique to estimate the probability of second
digits $d_2 = 0, \ldots, 9$. You may assume a function `second_digit()`
exists already.

<!-- answer-space: 1.6in -->

### Solution

The generation exercise, reading a different digit. Draw $F \sim U(0,1)$, set
$X = 10^F$, and tabulate the second digit. Code might look like this:

```{r echo = TRUE, eval = FALSE}
x <- 10^runif(n)
table(factor(second_digit(x), levels = 0:9)) / n
``` 

$M$ can be dropped entirely. It moves the decimal point and changes no digit,
which is why the leading digit depended only on $F$ in the first place. And
no new theory is needed: assuming $F$ uniform already fixes every digit, not
just the first.

**Rubric.** 1 pt for simulating $F \sim U(0,1)$ and reading the second digit.
Keeping $M$ is fine. No credit for "look up the second-digit law."
