---
title: "Module 1 Project: How Long Is a Perch?"
author: "Your Name, Your Uniqname"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(288942)
library(tidyverse)
```

\newcommand{\E}[1]{\text{E}\left(#1\right)}
\newcommand{\Var}[1]{\text{Var}\left(#1\right)}
\newcommand{\ind}{\mathbf{1}}

# The setting

The University of Helsinki recorded 159 fish caught in lake Laengelmavesi,
including 56 perch. We will model the **length of a perch** (the `length1`
measurement, in cm).

```{r}
fish <- read.csv("../../data/fish.csv")
perch <- filter(fish, species == 7)
ggplot(perch, aes(x = length1)) + geom_histogram(bins = 10)
```

Our model for perch length is supported on $[0, \theta]$, with $\theta = 50$
cm taken as the largest a perch could plausibly be:

**The model**: a **scaled Beta**: $X = 50 B$ with
$B \sim \text{Beta}(4, 4)$, so the density is
$$f(x) = \frac{1}{50} \, f_B(x / 50), \quad 0 \le x \le 50,$$
where $f_B$ is the Beta(4, 4) density. In general the
Beta$(\alpha, \beta)$ density on $[0, 1]$ is
$$f_B(u) = \frac{1}{\mathrm{B}(\alpha, \beta)}\,
  u^{\alpha - 1} (1 - u)^{\beta - 1}, \qquad 0 \le u \le 1,$$
where $\mathrm{B}(\alpha,\beta) = \Gamma(\alpha)\Gamma(\beta)/\Gamma(\alpha+\beta)$
is the **beta function** --- whatever constant makes the thing integrate to
one. It does not depend on $u$, which is why it drops out of most of what
follows and matters in exactly one place, C.2.

With $\alpha = \beta = 4$ the density is symmetric, with its peak at 25 cm:

```{r}
fmod <- function(x) { dbeta(x / 50, 4, 4) / 50 }
curve(fmod(x), from = -1, to = 55)
```

The shape parameter here is a *guess*, not a fact. Part C estimates it from
the data and Part D builds a confidence interval for it, so by the end you
will know how much of a guess it was.

Reminder: this document will be used on the exam. For any question make sure to display any requested output. If you use any AI assistance in implementing the R code, be prepared to explain how the implementation works.

# Part A: Monte Carlo integration

### A.1

Let $B$ be any random variable with CDF $F_B$ and density $f_B$, and let
$X = 50 B$. **Prove** that
$$F_X(x) = F_B(x / 50) \qquad \text{and} \qquad
f_X(x) = \frac{1}{50} f_B(x / 50).$$

Then **compute** the exact $F(20) = P(X \le 20)$ using `pbeta`.
We will treat this as the exact value for the rest of the project.

### A.2

Now estimate the same quantity by Monte Carlo, pretending A.1 were not
available:
$$F(20) = \int_0^{20} f(x) \, dx = \E{20 f(V)}, \quad V \sim U(0, 20).$$

Use 10,000 uniform draws. Report the estimate and a 95\% confidence
interval from `t.test`.

### A.3

Estimate the model's mean perch length,
$$\E{X} = \int_0^{50} x f(x) \, dx,$$
with a $U(0, 50)$ envelope and 10,000 draws. Report a 95\% confidence
interval and check against numerical integration:

```{r}
integrate(function(x) x * fmod(x), 0, 50)
```

The model is symmetric about 25, so $\E{X}$ is also available exactly in one
line. Give it, and set all three numbers side by side.

### A.4

Compare model to data: report the exact model probability $F(20)$ from A.1
next to the observed fraction of the 56 perch with `length1` $\le 20$.

### A.5

A joint quantity: two perch are caught independently, both following Model
B. Estimate the probability their **combined** length exceeds 70 cm,
$$P(X_1 + X_2 > 70)
= \int_0^{50}\!\!\int_0^{50} \ind(x_1 + x_2 > 70)\, f(x_1) f(x_2)
\, dx_1 \, dx_2,$$
using independent pairs $V_1, V_2 \sim U(0, 50)$ and the
multiply-and-divide recipe (careful: the joint envelope density is
$1/50^2$). Use 100,000 pairs and report a 95\% confidence interval. You
will check this answer in Part C.

# Part B: variance reduction

### B.1

Antithetic variables for the A.2 estimator. The model has its mode at 25, so
on $[0, 20]$ the density $f$ is strictly increasing (show this with a plot
or a one-line argument), and therefore $h(v) = 20 f(v)$ is monotone and the
pairs $(V, 20 - V)$ are negatively correlated.

Re-estimate $F(20)$ using 5,000 uniform draws $V_i$ paired with their
antithetic partners $20 - V_i$ (10,000 function evaluations total, same as
A.2). Report the estimate, the estimated variance of this estimator and of
the A.2 estimator, and the percent variance reduction.

### B.2

Importance sampling for $\E{X}$. The $U(0, 50)$ envelope in A.3 ignores
where $x f(x)$ puts its mass. Consider envelopes of the form $Y = 50 B'$
with $B' \sim \text{Beta}(\alpha, \beta)$, which by A.1 has density
$g(y) = \frac{1}{50}\,\texttt{dbeta}(y/50, \alpha, \beta)$ on $[0, 50]$
--- the uniform is the special case $\alpha = \beta = 1$.

Propose **two** candidate $(\alpha, \beta)$ pairs with small integer
values. Useful facts: a Beta($\alpha, \beta$) with $\alpha, \beta > 1$ has
its mode at $(\alpha - 1)/(\alpha + \beta - 2)$, and plotting candidates
against the target is the way to choose:

```{r eval = FALSE}
curve(x * fmod(x), from = 0, to = 50)
curve(K * dbeta(x / 50, alpha, beta) / 50, add = TRUE, col = "red")
## pick K so the two curves have comparable height

```

For each candidate (10,000 draws each), estimate $\E{X}$ and the variance
of the estimator, and compare with the uniform envelope of A.3.

One choice of small integers $(\alpha, \beta)$ turns out to be **perfect**:
every weighted value $y \, f(y) / g(y)$ is the same number. Find it (the
mode-matching hint leads straight there), and verify by printing the
`range()` of your weighted values. Compare session exercise 3(e).

### B.3

Using the estimated variance of your best **non-perfect** envelope from
B.2, compute the number of draws needed for a 95\% confidence interval for
$\E{X}$ with width at most $0.1$ cm. Then run it once with that $n$ and
report the realized interval and its width.

# Part C: Monte Carlo for estimation

Sampling from the model is one line --- scale draws from `rbeta`:

```{r}
rmod <- function(n) { 50 * rbeta(n, 4, 4) }
```

So far $\text{Beta}(4,4)$ was handed to you. Now we estimate it. Keep the
upper limit $\theta = 50$ fixed and keep the symmetry, so the model has a
**single unknown parameter** $a$:
$$X = 50 B, \qquad B \sim \text{Beta}(a, a).$$
Our working model is the case $a = 4$.

### C.1

**Method of moments.** Using
$\Var{B} = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$, show that
$$\Var{X} = \frac{625}{2a + 1},$$
and hence that setting $\Var{X} = S^2$ and solving gives
$$\tilde a = \frac{1}{2}\left(\frac{625}{S^2} - 1\right).$$
Compute $\tilde a$ for the 56 perch.

Note there is no second equation to solve: $\E{X} = 25$ for every $a$, so
the mean carries no information about the parameter and the variance has to
do all the work.

### C.2

**Maximum likelihood.** Here the beta function finally earns its keep.
Setting $\alpha = \beta = a$ in the density from the setting,
$$f_B(u) = \frac{1}{\mathrm{B}(a, a)}\, u^{a - 1} (1 - u)^{a - 1},$$
and dividing by 50 for the scaling (A.1), gives the model's density. With
$u_i = x_i / 50$, show that the log-likelihood is
$$\ell(a) = -n \log 50 - n \log \mathrm{B}(a,a)
 + (a - 1) \sum_{i=1}^n \left[\log u_i + \log(1 - u_i)\right],$$
where $\mathrm{B}$ is the beta function (`beta()` in R, `lbeta()` for its
log). Conclude that the data enters **only** through the single number
$T = \sum_i [\log u_i + \log(1-u_i)]$.

Unlike the $U(0,\theta)$ maximum in video 3, this likelihood is smooth in $a$, so
here you may differentiate. Either solve $\ell'(a) = 0$ numerically with
`uniroot` (you will need $\frac{d}{da}\log\Gamma(a) = \psi(a)$, which is
`digamma()` in R), or maximize $\ell$ directly with `optimize`. Report
$\hat a$ and compare it with $\tilde a$ from C.1.

### C.3

Now pretend we do not know $F(20)$ and must estimate it from a sample of
$n = 56$. Two estimators:

- **ECDF**: $\hat F(20) = \frac{1}{56} \sum_i \ind(X_i \le 20)$
- **Plug-in**: estimate $a$ by the method of moments from C.1, then use the
  exact CDF form from A.1: `pbeta(20/50, atilde, atilde)`.

Simulate the sampling distribution of each: 1,000 replications of samples
of size 56 from the model ($a = 4$). Using the exact $F(20)$ from A.1,
estimate the **bias** and **MSE** of each, with a 99\% confidence interval,
and show a side-by-side boxplot.

### C.4

Repeat the comparison at $n = 14$ and $n = 224$. Present bias, variance,
and MSE for both estimators at all three sample sizes in one table.

### C.5

The $t$ confidence interval is exact only for Normal data, and the model is
not Normal. Estimate its true coverage: simulate 10,000 samples of size 56
from the model, compute the nominal 95\% `t.test` interval for $\E{X}$ in
each, and report the proportion covering the true value (from A.3), with a
99\% confidence interval for that proportion (`binom.test`).

Then check your A.5 answer by direct sampling: estimate $P(X_1 + X_2 > 70)$
as the mean of indicators over 100,000 `rmod` pairs, with a 99\%
`binom.test` interval.

# Part D: Monte Carlo for testing

Part C estimated $a$. Now we **test** it. Throughout, $H_0$ says the perch
follow our working model,
$$H_0: \; X = 50B, \quad B \sim \text{Beta}(4, 4),$$
with $n = 56$, and the alternatives are other values of $a$.

Two statistics:

- **Moment statistic** $T_S = S^2$, the sample variance. Under the model the
  variance is $625/(2a + 1)$ from C.1, so $T_S$ is a direct read on $a$.
- **Distance statistic** $T_K$, the largest gap between the model CDF and
  the empirical CDF --- `ks.test(x/50, "pbeta", 4, 4)$statistic`. It looks
  at the whole distribution rather than at $a$ in particular.

Neither has a usable table here, which is why we simulate.

### D.1

Simulate the null distribution of each statistic: 10,000 samples of size 56
from the model with $a = 4$.

$T_S$ can be wrong in either direction, so use a **two-sided** region,
rejecting when $T_S$ falls below the 2.5\% quantile or above the 97.5\%
quantile. $T_K$ is a distance, so reject only above its 95\% quantile.
Report all three cutoffs.

### D.2

With **fresh** simulations under $H_0$ (10,000 new replications), estimate
the empirical Type I error rate of each region, each with a `binom.test`
confidence interval.

### D.3

Power curves. For each $a \in \{2, 2.5, 3, 3.5, \ldots, 8\}$, simulate
1,500 samples of size 56 from $50 \cdot \text{Beta}(a, a)$ and estimate the
power of both tests, using the cutoffs from D.1.

Plot the two power curves on one set of axes, with horizontal reference
lines at $0.05$ and $0.80$.

### D.4

**(a)** Test $H_0: a = 4$ on the 56 perch lengths using $T_S$. Report the
observed $S^2$, its $p$-value in the $(1 + \#)/(B + 1)$ form, and whether
the test rejects at the $0.05$ level.

**(b)** Invert that test to get a confidence interval for $a$. For each
candidate $a_0$ on a grid from $1.5$ to $8$ in steps of $0.1$:

- simulate the null distribution of $T_S$ under
  $50 \cdot \text{Beta}(a_0, a_0)$, 2,000 replications;
- keep $a_0$ if the observed $S^2$ falls between the $2.5\%$ and $97.5\%$
  quantiles of that distribution.

The kept values form a 95\% confidence interval for $a$. Report its
endpoints, and plot the grid against the $p$-value with $\tilde a$ from C.1
and the working value $a = 4$ marked.
