---
title: "Module 1 Project: How Long Is a Perch?"
author: "Your Name, Your Uniqname"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(288942)
library(tidyverse)
```

\newcommand{\E}[1]{\text{E}\left(#1\right)}
\newcommand{\Var}[1]{\text{Var}\left(#1\right)}
\newcommand{\ind}{\mathbf{1}}

# The setting

The University of Helsinki recorded 159 fish caught in lake Laengelmavesi,
including 56 perch. We will model the **length of a perch** (the `length1`
measurement, in cm).

```{r}
fish <- read.csv("../../data/fish.csv")
perch <- filter(fish, species == 7)
ggplot(perch, aes(x = length1)) + geom_histogram(bins = 10)
```

Two candidate models for perch length, both supported on $[0, \theta]$ with
$\theta = 50$:

**Model U** (crude): $X \sim U(0, \theta)$.

**Model B** (better): a **scaled Beta**: $X = 50 B$ with
$B \sim \text{Beta}(5, 2)$, so the density is
$$f(x) = \frac{1}{50} \, f_B(x / 50), \quad 0 \le x \le 50,$$
where $f_B$ is the Beta(5, 2) density. This puts the peak at 40 cm:

```{r}
fmod <- function(x) { dbeta(x / 50, 5, 2) / 50 }
curve(fmod(x), from = -1, to = 55)
```

Parts A and B use only the multiply-and-divide recipe from the video: for
any density $g$ positive on the region,
$$\int_a^b h(x) \, dx = \E{\frac{h(Y)}{g(Y)}}, \quad Y \sim g.$$
Part C starts sampling from Model B directly.

You will build one analysis of these models across the module, adding a part
after each video. **The module exam asks questions about what you did and
found here** --- as you work, make sure you could explain each choice and
each result, because the exam will ask you to.

Submit one knitted document with all code visible.

# Part A: Monte Carlo integration (after video 1) (10 pts)

### A.1 (3 pts)

Let $B$ be any random variable with CDF $F_B$ and density $f_B$, and let
$X = 50 B$. **Prove** that
$$F_X(x) = F_B(x / 50) \qquad \text{and} \qquad
f_X(x) = \frac{1}{50} f_B(x / 50).$$
(Start from $F_X(x) = P(50 B \le x)$; get the density from the CDF.)

Then **compute** the exact $F(30) = P(X \le 30)$ for Model B using `pbeta`.
We will treat this as the exact value for the rest of the project.

### A.2 (2 pts)

Now estimate the same quantity by Monte Carlo, pretending A.1 were not
available:
$$F(30) = \int_0^{30} f(x) \, dx = \E{30 f(V)}, \quad V \sim U(0, 30).$$

Use 10,000 uniform draws. Report the estimate and a 95\% confidence
interval from `t.test`. Does the interval cover the exact value?

### A.3 (2 pts)

Estimate the model's mean perch length,
$$\E{X} = \int_0^{50} x f(x) \, dx,$$
with a $U(0, 50)$ envelope and 10,000 draws. Report a 95\% confidence
interval and check against numerical integration:

```{r}
integrate(function(x) x * fmod(x), 0, 50)
```

### A.4 (1 pt)

Compare model to data: report the exact model probability $F(30)$ from A.1
next to the observed fraction of the 56 perch with `length1` $\le 30$.
Keep this comparison in mind --- we return to it in Part D.

### A.5 (2 pts)

A joint quantity: two perch are caught independently, both following Model
B. Estimate the probability their **combined** length exceeds 90 cm,
$$P(X_1 + X_2 > 90)
= \int_0^{50}\!\!\int_0^{50} \ind(x_1 + x_2 > 90)\, f(x_1) f(x_2)
\, dx_1 \, dx_2,$$
using independent pairs $V_1, V_2 \sim U(0, 50)$ and the
multiply-and-divide recipe (careful: the joint envelope density is
$1/50^2$). Use 100,000 pairs and report a 95\% confidence interval. You
will check this answer in Part C.

# Part B: variance reduction (after video 2) (10 pts)

### B.1 (3 pts)

Antithetic variables for the A.2 estimator. On $[0, 30]$ the density $f$ is
strictly increasing (show this with a plot or a one-line argument --- where
is the mode?), so $h(v) = 30 f(v)$ is monotone and the pairs $(V, 30 - V)$
are negatively correlated.

Re-estimate $F(30)$ using 5,000 uniform draws $V_i$ paired with their
antithetic partners $30 - V_i$ (10,000 function evaluations total, same as
A.2). Report the estimate, the estimated variance of this estimator and of
the A.2 estimator, and the percent variance reduction.

### B.2 (5 pts)

Importance sampling for $\E{X}$. The $U(0, 50)$ envelope in A.3 ignores
where $x f(x)$ puts its mass. Consider envelopes of the form $Y = 50 B'$
with $B' \sim \text{Beta}(\alpha, \beta)$, which by A.1 has density
$g(y) = \frac{1}{50}\,\texttt{dbeta}(y/50, \alpha, \beta)$ on $[0, 50]$
--- the uniform is the special case $\alpha = \beta = 1$.

Propose **two** candidate $(\alpha, \beta)$ pairs with small integer
values. Useful facts: a Beta($\alpha, \beta$) with $\alpha, \beta > 1$ has
its mode at $(\alpha - 1)/(\alpha + \beta - 2)$, and plotting candidates
against the target is the way to choose:

```{r eval = FALSE}
curve(x * fmod(x), from = 0, to = 50)
curve(K * dbeta(x / 50, alpha, beta) / 50, add = TRUE, col = "red")
## pick K so the two curves have comparable height
```

For each candidate (10,000 draws each), estimate $\E{X}$ and the variance
of the estimator, and compare with the uniform envelope of A.3.

One choice of small integers $(\alpha, \beta)$ turns out to be **perfect**:
every weighted value $y \, f(y) / g(y)$ is the same number. Find it (the
mode-matching hint leads straight there), and verify by printing the
`range()` of your weighted values. Compare session exercise 3(e).

### B.3 (2 pts)

Using the estimated variance of your best **non-perfect** envelope from
B.2, compute the number of draws needed for a 95\% confidence interval for
$\E{X}$ with width at most $0.1$ cm. Then run it once with that $n$ and
report the realized interval and its width.

# Part C: Monte Carlo for estimation (after video 3) (10 pts)

Sampling from Model B is one line --- scale draws from `rbeta`:

```{r}
rmod <- function(n) { 50 * rbeta(n, 5, 2) }
```

Now pretend we do *not* know $F(30)$ and must estimate it from a sample of
$n = 56$ perch, the size of our real sample. Two estimators:

- **ECDF**: $\hat F(30) = \frac{1}{56} \sum_i \ind(X_i \le 30)$
- **Plug-in**: fit the Beta parameters by the **method of moments** and use
  the exact CDF form from A.1. With $\hat m = \bar X / 50$ and
  $\hat v = S^2 / 50^2$,
  $$\hat k = \frac{\hat m (1 - \hat m)}{\hat v} - 1, \qquad
  \hat\alpha = \hat m \hat k, \qquad \hat\beta = (1 - \hat m) \hat k,$$
  and the estimator is `pbeta(30/50, alphahat, betahat)`.

### C.1 (2 pts)

Simulate the sampling distribution of the ECDF estimator: 1,000 replications
of samples of size 56 from Model B. Using the exact $F(30)$ from A.1,
estimate the **bias** and **MSE**, each with a 99\% confidence interval.

### C.2 (3 pts)

Repeat C.1 for the plug-in estimator. Compare the two on bias and MSE, with
a side-by-side boxplot of the two sampling distributions. Which do you
prefer at $n = 56$?

### C.3 (2 pts)

Repeat the comparison at $n = 14$ and $n = 224$. Present bias, variance,
and MSE for both estimators at all three sample sizes in one table. Does
the preferred estimator change with $n$?

### C.4 (2 pts)

The $t$ confidence interval is exact only for Normal data, and Model B is
not Normal. Estimate its true coverage: simulate 10,000 samples of size 56
from Model B, compute the nominal 95\% `t.test` interval for $\E{X}$ in
each, and report the proportion covering the true value (from A.3), with a
99\% confidence interval for that proportion (`binom.test`).

### C.5 (1 pt)

Check your A.5 answer with direct sampling: estimate $P(X_1 + X_2 > 90)$ as
the mean of indicators over 100,000 `rmod` pairs, with a 99\% `binom.test`
interval. Do the two approaches agree?

# Part D: Monte Carlo for testing (after video 4) (10 pts)

Under Model U, two estimators of $\theta$ from earlier in the course double
as test statistics:

- Method of moments: $\tilde\theta = 2 \bar X$
- Maximum likelihood: $\hat\theta = \max_i X_i$

We test $H_0: \theta = 50$ against $H_1: \theta = 51$ with $n = 56$.

### D.1 (3 pts)

Simulate the null ($\theta = 50$) and alternative ($\theta = 51$)
distributions of both statistics, 10,000 replications each. For each
statistic, find a rejection region of the form $T > c$ with
$P(T > c \mid H_0) \le 0.05$, and estimate its power against $\theta = 51$.

### D.2 (2 pts)

Check your regions: with **fresh** simulations under $H_0$ (10,000 new
replications), estimate the empirical Type I error rate of each rejection
region, with a confidence interval. Are both at or below 0.05?

### D.3 (3 pts)

Power curves: estimate the power of both tests for each
$\theta \in \{51, 52, \ldots, 80\}$ (1,000 replications per $\theta$ is
enough). Add a **third** competitor: under Model U, $\E{X} = \theta/2$, so
testing $\theta = 50$ is testing $\mu_X = 25$ --- use
`t.test(x, mu = 25, alternative = "greater")` and reject when its p-value
is below 0.05. Plot all three curves on one set of axes and report how
large $\theta$ must be for each test to reach 80\% power.

### D.4 (2 pts)

Back to the data: carry out the test from D.1 (your preferred statistic) on
the 56 perch lengths and report the outcome. Then estimate $P(X \le 30)$
for the perch using the estimator you preferred in C.2, and set it beside
the A.4 comparison. State your conclusions about the two models in plain
language.
