---
title: "Module 1 Project: How Long Is a Perch?"
author: "Your Name, Your Uniqname"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(288942)
library(tidyverse)
```

\newcommand{\E}[1]{\text{E}\left(#1\right)}
\newcommand{\Var}[1]{\text{Var}\left(#1\right)}
\newcommand{\ind}{\mathbf{1}}

# The setting

The University of Helsinki recorded 159 fish caught in lake Laengelmavesi,
including 56 perch. We will model the **length of a perch** (the `length1`
measurement, in cm).

```{r}
fish <- read.csv("../../data/fish.csv")
perch <- filter(fish, species == 7)
ggplot(perch, aes(x = length1)) + geom_histogram(bins = 10)
```

Our model for perch length is supported on $[0, \theta]$, with $\theta = 50$
cm taken as the largest a perch could plausibly be:

**Model B**: a **scaled Beta**: $X = 50 B$ with
$B \sim \text{Beta}(4, 4)$, so the density is
$$f(x) = \frac{1}{50} \, f_B(x / 50), \quad 0 \le x \le 50,$$
where $f_B$ is the Beta(4, 4) density. In general the
Beta$(\alpha, \beta)$ density on $[0, 1]$ is
$$f_B(u) = \frac{1}{\mathcal{B}(\alpha, \beta)}\,
  u^{\alpha - 1} (1 - u)^{\beta - 1}, \qquad 0 \le u \le 1,$$
where $\mathcal{B}(\alpha,\beta) = \Gamma(\alpha)\Gamma(\beta)/\Gamma(\alpha+\beta)$
is the **beta function** --- whatever constant makes the thing integrate to
one. It does not depend on $u$, which is why it drops out of most of what
follows and matters in exactly one place, C.2.

With $\alpha = \beta = 4$ the density is symmetric, with its peak at 25 cm:

```{r}
fmod <- function(x) { dbeta(x / 50, 4, 4) / 50 }
curve(fmod(x), from = -1, to = 55)
```

Model B is a *guess*, not a fact. Its shape and its upper limit are both
things we could be wrong about, and in **Part D** you will build a
confidence region for them and find out whether this guess survives.

Reminder: this document will be used on the exam. For any question make sure to display any requested output. If you use any AI assistance in implementing the R code, be prepared to explain how the implementation works.

# Part A: Monte Carlo integration

### A.1

Let $B$ be any random variable with CDF $F_B$ and density $f_B$, and let
$X = 50 B$. **Prove** that
$$F_X(x) = F_B(x / 50) \qquad \text{and} \qquad
f_X(x) = \frac{1}{50} f_B(x / 50).$$

Then **compute** the exact $F(20) = P(X \le 20)$ for Model B using `pbeta`.
We will treat this as the exact value for the rest of the project.

### A.2

Now estimate the same quantity by Monte Carlo, pretending A.1 were not
available:
$$F(20) = \int_0^{20} f(x) \, dx = \E{20 f(V)}, \quad V \sim U(0, 20).$$

Use 10,000 uniform draws. Report the estimate and a 95\% confidence
interval from `t.test`.

### A.3

Estimate the model's mean perch length,
$$\E{X} = \int_0^{50} x f(x) \, dx,$$
with a $U(0, 50)$ envelope and 10,000 draws. Report a 95\% confidence
interval and check against numerical integration:

```{r}
integrate(function(x) x * fmod(x), 0, 50)
```

Model B is symmetric about 25, so you can also write down $\E{X}$ exactly
in one line. Do so, and say which of your three numbers you trust most and
why.

### A.4

Compare model to data: report the exact model probability $F(20)$ from A.1
next to the observed fraction of the 56 perch with `length1` $\le 20$.
Keep this comparison in mind --- we return to it in Part D.

### A.5

A joint quantity: two perch are caught independently, both following Model
B. Estimate the probability their **combined** length exceeds 70 cm,
$$P(X_1 + X_2 > 70)
= \int_0^{50}\!\!\int_0^{50} \ind(x_1 + x_2 > 70)\, f(x_1) f(x_2)
\, dx_1 \, dx_2,$$
using independent pairs $V_1, V_2 \sim U(0, 50)$ and the
multiply-and-divide recipe (careful: the joint envelope density is
$1/50^2$). Use 100,000 pairs and report a 95\% confidence interval. You
will check this answer in Part C.

# Part B: variance reduction

### B.1

Antithetic variables for the A.2 estimator. Model B has its mode at 25, so
on $[0, 20]$ the density $f$ is strictly increasing (show this with a plot
or a one-line argument), and therefore $h(v) = 20 f(v)$ is monotone and the
pairs $(V, 20 - V)$ are negatively correlated.

Re-estimate $F(20)$ using 5,000 uniform draws $V_i$ paired with their
antithetic partners $20 - V_i$ (10,000 function evaluations total, same as
A.2). Report the estimate, the estimated variance of this estimator and of
the A.2 estimator, and the percent variance reduction.

Then say what would have gone wrong had we kept the interval at $[0, 30]$.

### B.2

Importance sampling for $\E{X}$. The $U(0, 50)$ envelope in A.3 ignores
where $x f(x)$ puts its mass. Consider envelopes of the form $Y = 50 B'$
with $B' \sim \text{Beta}(\alpha, \beta)$, which by A.1 has density
$g(y) = \frac{1}{50}\,\texttt{dbeta}(y/50, \alpha, \beta)$ on $[0, 50]$
--- the uniform is the special case $\alpha = \beta = 1$.

Propose **two** candidate $(\alpha, \beta)$ pairs with small integer
values. Useful facts: a Beta($\alpha, \beta$) with $\alpha, \beta > 1$ has
its mode at $(\alpha - 1)/(\alpha + \beta - 2)$, and plotting candidates
against the target is the way to choose:

```{r eval = FALSE}
curve(x * fmod(x), from = 0, to = 50)
curve(K * dbeta(x / 50, alpha, beta) / 50, add = TRUE, col = "red")
## pick K so the two curves have comparable height

```

For each candidate (10,000 draws each), estimate $\E{X}$ and the variance
of the estimator, and compare with the uniform envelope of A.3.

One choice of small integers $(\alpha, \beta)$ turns out to be **perfect**:
every weighted value $y \, f(y) / g(y)$ is the same number. Find it (the
mode-matching hint leads straight there), and verify by printing the
`range()` of your weighted values. Compare session exercise 3(e).

### B.3

Using the estimated variance of your best **non-perfect** envelope from
B.2, compute the number of draws needed for a 95\% confidence interval for
$\E{X}$ with width at most $0.1$ cm. Then run it once with that $n$ and
report the realized interval and its width.

# Part C: Monte Carlo for estimation

Sampling from Model B is one line --- scale draws from `rbeta`:

```{r}
rmod <- function(n) { 50 * rbeta(n, 4, 4) }
```

So far $\text{Beta}(4,4)$ was handed to you. Now we estimate it. Keep the
upper limit $\theta = 50$ fixed and keep the symmetry, so Model B has a
**single unknown parameter** $a$:
$$X = 50 B, \qquad B \sim \text{Beta}(a, a).$$
Our working model is the case $a = 4$.

### C.1

**Method of moments.** Using
$\Var{B} = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$, show that
$$\Var{X} = \frac{625}{2a + 1},$$
and hence that setting $\Var{X} = S^2$ and solving gives
$$\tilde a = \frac{1}{2}\left(\frac{625}{S^2} - 1\right).$$
Compute $\tilde a$ for the 56 perch.

(Why is there no equation for the *mean* here? Say what $\E{X}$ is under
this model and why it carries no information about $a$.)

### C.2

**Maximum likelihood.** Here the beta function finally earns its keep.
Setting $\alpha = \beta = a$ in the density from the setting,
$$f_B(u) = \frac{1}{\mathcal{B}(a, a)}\, u^{a - 1} (1 - u)^{a - 1},$$
and dividing by 50 for the scaling (A.1), gives Model B's density. With
$u_i = x_i / 50$, show that the log-likelihood is
$$\ell(a) = -n \log 50 - n \log \mathcal{B}(a,a)
 + (a - 1) \sum_{i=1}^n \left[\log u_i + \log(1 - u_i)\right],$$
where $\mathcal{B}$ is the beta function (`beta()` in R, `lbeta()` for its
log). Conclude that the data enters **only** through the single number
$T = \sum_i [\log u_i + \log(1-u_i)]$.

Unlike the $U(0,\theta)$ maximum in video 3, this likelihood is smooth in $a$, so
here you may differentiate. Either solve $\ell'(a) = 0$ numerically with
`uniroot` (you will need $\frac{d}{da}\log\Gamma(a) = \psi(a)$, which is
`digamma()` in R), or maximize $\ell$ directly with `optimize`. Report
$\hat a$ and compare it with $\tilde a$ from C.1.

### C.3

Now pretend we do not know $F(20)$ and must estimate it from a sample of
$n = 56$. Two estimators:

- **ECDF**: $\hat F(20) = \frac{1}{56} \sum_i \ind(X_i \le 20)$
- **Plug-in**: estimate $a$ by the method of moments from C.1, then use the
  exact CDF form from A.1: `pbeta(20/50, atilde, atilde)`.

Simulate the sampling distribution of each: 1,000 replications of samples
of size 56 from Model B ($a = 4$). Using the exact $F(20)$ from A.1,
estimate the **bias** and **MSE** of each, with a 99\% confidence interval,
and show a side-by-side boxplot. Which do you prefer at $n = 56$?

### C.4

Repeat the comparison at $n = 14$ and $n = 224$. Present bias, variance,
and MSE for both estimators at all three sample sizes in one table. Does
the preferred estimator change with $n$?

### C.5

The $t$ confidence interval is exact only for Normal data, and Model B is
not Normal. Estimate its true coverage: simulate 10,000 samples of size 56
from Model B, compute the nominal 95\% `t.test` interval for $\E{X}$ in
each, and report the proportion covering the true value (from A.3), with a
99\% confidence interval for that proportion (`binom.test`).

Then check your A.5 answer by direct sampling: estimate $P(X_1 + X_2 > 70)$
as the mean of indicators over 100,000 `rmod` pairs, with a 99\%
`binom.test` interval. Do the two approaches agree?

# Part D: Monte Carlo for testing

Part C estimated $a$. Now we **test** it. Throughout, $H_0$ says the perch
follow our working model,
$$H_0: \; X = 50B, \quad B \sim \text{Beta}(4, 4),$$
and $n = 56$ as before.

Two statistics, aimed at rather different things:

- **Moment statistic** $T_S = S^2$, the sample variance. Under Model B the
  variance is $625/(2a+1)$ from C.1, so $T_S$ is a direct read on $a$ and on
  nothing else.
- **Distance statistic** $T_K$, the largest gap between the model CDF and
  the empirical CDF --- `ks.test(x/50, "pbeta", 4, 4)$statistic`. It looks at
  the whole distribution and is aimed at nothing in particular.

Neither has a usable table here, which is why we simulate.

### D.1

Simulate the null distribution of each statistic: 10,000 samples of size 56
from Model B with $a = 4$.

$T_S$ can be wrong in either direction, so use a **two-sided** region,
rejecting when $T_S$ is below the 2.5\% quantile or above the 97.5\%
quantile. $T_K$ is a distance, so large values are the only evidence against
$H_0$: reject above its 95\% quantile. Report all three cutoffs.

### D.2

Check your regions: with **fresh** simulations under $H_0$ (10,000 new
replications), estimate the empirical Type I error rate of each region, with
a `binom.test` confidence interval. Are both near 0.05? Explain in one
sentence why the simulations from D.1 could not have told you this.

### D.3

Now compare the two tests, at the same size, against **two** different ways
of being wrong.

The first changes the **spread** and nothing else: $a = 6$, still symmetric.

The second changes the **shape** while holding the spread fixed at the
Model B value. Use the mode/concentration parameterization: a Beta with mode
$m$ (as a fraction of $\theta$) and concentration $c = \alpha + \beta$ has
$$\alpha = 1 + m(c - 2), \qquad \beta = 1 + (1 - m)(c - 2),$$
and this helper picks $c$ so the standard deviation comes out to `s`:

```{r}
beta_from_mode <- function(theta, m, s) {
  g <- function(cc) {
    a <- 1 + m * (cc - 2); b <- 1 + (1 - m) * (cc - 2)
    theta * sqrt(a * b / ((a + b)^2 * (a + b + 1))) - s
  }
  cc <- uniroot(g, c(2.001, 500))$root
  c(alpha = 1 + m * (cc - 2), beta = 1 + (1 - m) * (cc - 2))
}
```

Call it with `theta = 50`, `m = 0.28`, and `s` set to Model B's own standard
deviation, so this alternative has **exactly the right variance** and only
the shape is wrong.

Estimate the power of both tests against both alternatives (2,000
replications each) and put the four numbers in one table.

### D.4

The table should not be a clean sweep --- one test wins one row and loses
the other, by a wide margin both times.

Explain it. Why is $T_S$ nearly powerless against the second alternative,
and what is its power there close to? Why does $T_K$ do comparatively badly
on the first? State the general lesson in one sentence, in the form "a test
aimed at ... ".

Then apply both tests to the 56 perch: report a $p$-value for $H_0$ from
each, using the $(1 + \#)/(B+1)$ form. Do they agree about Model B?

Finally, say which of the two you would want in D.5, where the question is
not "is $a$ equal to 4" but "is this model acceptable at all".

### D.5

**Back to Model B: is it any good?** Throughout Parts A--C we fixed
$\theta = 50$ and assumed symmetry. Now let both go, and build a
**confidence region** for them by inverting a test.

Reuse the $(\theta, m)$ parameterization from D.3 and the `beta_from_mode`
helper, now calling it with `s = sd(perch$length1)` so that every candidate
model is at least the right width. That leaves $\theta$ and $m$ as the two
things actually being tested, and a grid over them is a two-dimensional
search.

Use $T_K$ as the statistic, for the reason you gave at the end of D.4.

**(a)** Write a function that, for a given $(\theta, m)$, gets
$(\alpha,\beta)$ from `beta_from_mode`, computes the observed statistic on
`perch$length1 / theta`, simulates its null distribution with `rbeta`, and
returns the $(1 + \#)/(B + 1)$ $p$-value. Use $B = 200$. Reject
$(\theta, m)$ when that $p$-value is at or below 0.05.

**(b)** Any $\theta$ below `max(perch$length1)` is impossible --- say why in
one sentence, and exclude it.

**(c)** Run the test over a grid of $\theta$ from 42 to 64 and $m$ from 0.25
to 0.75. Plot the accepted set. Keep the grid coarse enough to finish in
under a minute, then refine it if you like. If the accepted set runs off the
edge of your grid, say so --- that is a result, not a bug.

**(d)** Report the range of $\theta$ in your region, the range of $m$, and
the range of the **mode in centimetres**, $\theta \times m$. Which of the
three does the data determine most sharply? For any that it does **not**
determine, say what that means in terms of perch: what question are these
56 fish simply unable to answer?

**(e)** Is the grid point $(\theta, m) = (50, 0.5)$ inside your region?

First print `beta_from_mode(50, 0.5, sd(perch$length1))` and set it beside
$\tilde a$ from C.1. They agree **exactly**. Show algebraically why, in one
line --- at $m = 0.5$ the helper is solving the same equation C.1 did.

So this grid point is not the $\text{Beta}(4,4)$ of Parts A--C; it is the
symmetric Beta that matches the observed spread. Say what its being inside
the region does, and does not, entitle you to conclude about Model B.
